### 2024-04-22 abstract.tex
Added abstract

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts

\begin{document}

\title{Ensuring Fairness with Transparent Auditing of Ethical Bias in AI Systems\\
\thanks{Identify applicable funding agency here. If none, delete this.}
}

\author{\IEEEauthorblockN{1\textsuperscript{st} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
}

\maketitle

\begin{abstract}
With the rapid advancement in the field of AI, there is a growing trend to integrate AI into the information flow of the decision making processes. However, the deployment of AI algorithms in this respect raises ethical considerations, particularly, regarding fairness. AI systems may exhibit biases, caused by various sources, that can potentially lead decision makers to derive unfair conclusions. For example, in 2016, journalists at ProPublica discovered that COMPAS, the algorithm system used by the American justice system to evaluate recidivism, displayed unfair treatment to different racial groups, favoring the majority groups while harming the minority groups; specifically, it violates a fairness measure called equalized odds. In recent years, researchers have devised a number of measures designed to evaluate the fairness of AI systems. We propose here a framework for auditing fairness of AI systems by including a third party auditor and an AI system provider and create a tool to facilitate the examining of such systems. It's pivotal for the integrity of the audits that the framework includes an independent and trusted arbiter for their objectivity and accountability; third party auditors are often further necessary for their specialized expertise in the relevant domains such as medicine and law. Auditors, equipped with our tool, can thoroughly review the AI systems for bias and fairness violations. The tool is open-sourced, easily accessible, and publicly available. Unlike traditional AI systems, we advocate a transparent white-box and statistics based approach. It can be utilized by third party auditors, model makers, or the general public as a reference point when judging the fairness criterion of AI systems.
\end{abstract}

\begin{IEEEkeywords}
keyword, keyword
\end{IEEEkeywords}

\end{document}

### 2024-04-22 abstract.tex
Wrap text
Revised abstract

@@
 \begin{abstract}
-With the rapid advancement in the field of AI, there is a growing trend to integrate AI into the information flow of the decision making processes.
-However, the deployment of AI algorithms in this respect raises ethical considerations, particularly, regarding fairness.
-AI systems may exhibit biases, caused by various sources, that can potentially lead decision makers to derive unfair conclusions.
-For example, in 2016, journalists at ProPublica discovered that COMPAS, the algorithm system used by the American justice system to evaluate recidivism, displayed unfair treatment to different racial groups, favoring the majority groups while harming the minority groups; specifically, it violates a fairness measure called equalized odds.
-In recent years, researchers have devised a number of measures designed to evaluate the fairness of AI systems.
-We propose here a framework for auditing fairness of AI systems by including a third party auditor and an AI system provider and create a tool to facilitate the examining of such systems.
-It's pivotal for the integrity of the audits that the framework includes an independent and trusted arbiter for their objectivity and accountability; third party auditors are often further necessary for their specialized expertise in the relevant domains such as medicine and law.
-Auditors, equipped with our tool, can thoroughly review the AI systems for bias and fairness violations.
-The tool is open-sourced, easily accessible, and publicly available.
-Unlike traditional AI systems, we advocate a transparent white-box and statistics based approach.
-It can be utilized by third party auditors, model makers, or the general public as a reference point when judging the fairness criterion of AI systems.
+With the rapid advancement of AI, there is a growing trend to integrate AI into decision-making processes.
+However, AI systems may exhibit biases that lead decision makers to draw unfair conclusions.
+Notably, the COMPAS system used in the American justice system to evaluate recidivism was found to favor racial majority groups; specifically, it violates a fairness standard called equalized odds.
+Various measures have been proposed to assess AI fairness.
+We present a framework for auditing AI fairness, involving third-party auditors and AI system provider, and we have created a tool to facilitate systematic examination of AI systems.
+Inclusion of independent and trusted auditors is pivotal for objectivity and accountability; third parties are often further necessary for their specialized expertise in relevant domains.
+Auditors, equipped with our tool, can thoroughly review AI systems for bias and fairness violations.
+The tool is open-sourced and publicly available.
+Unlike traditional AI systems, we advocate a transparent white-box and statistics-based approach.
+It can be utilized by third party auditors, AI developers, or the general public for reference when judging the fairness criterion of AI systems.
 \end{abstract}

### 2024-04-24 abstract.tex
Added intro first two paragraph

@@
 
-\begin{IEEEkeywords}
-keyword, keyword
-\end{IEEEkeywords}
+\section{Introduction}
+In recent years, the accelerating pace of artificial intelligence (AI) technologies have revolutionized numerous fields.
+In healthcare, AI-driven diagnostic systems streamline disease identification, while in finance, automated trading algorithms analyze market trends to execute optimal trades swiftly.
+This remarkable progress has reached diverse industries,sprouting out various applications for different purposes.
+
+One of its most profound impact is how AI has transformed decision-making processes across sectors.
+By harnessing vast amounts of data and employing advanced algorithms, AI empowers organizations to make more informed and strategic decisions.
+From assessing job applicants to determining school admissions, AI-driven insights offer efficient and analytical advantages.
+
+\section{Ease of Use}
+bias sources example (COMPAS)
+fairness is subjective, based on culture and often not quantitative
+for AI system we use quantitative fairness measures
+transparency and objectivity example : COMPAS
+third party needed for auditing example : propublica like entity can use our stuff; we're not replacing but supplement; domain knowledge
+we propose a framework
 

### 2024-04-24 abstract.tex
Added compas intro

@@
 
+AI system, however, may be biased.
+Factors such as inherent biases in original data sets or flaws in algorithm designs could contribute to bias within AI systems.
+When left unchecked, the ramifications of such biases extend far beyond mere inaccuracies; it can lead to devastating consequences, such as amplifying systemic injustices, perpetuating group discrimination, and exacerbating societal inequalities.
+
+Correctional Offender Management Profiling for Alternative Sanctions (COMPAS) is a tool for predicting recidivism---the tendency of criminals to reoffend.
+It is used in the criminal justice system in multiple states in the United States.
+In 2016, it was discovered in an investigation by the journalists at ProPublica that the COMPAS system is, in fact, unfair towards minority and disadvantaged groups.
+
 \section{Ease of Use}
-bias sources example (COMPAS)
 fairness is subjective, based on culture and often not quantitative

### 2024-04-25 abstract.tex
added introduction

@@
 
-\section{Ease of Use}
-fairness is subjective, based on culture and often not quantitative
-for AI system we use quantitative fairness measures
-transparency and objectivity example : COMPAS
-third party needed for auditing example : propublica like entity can use our stuff; we're not replacing but supplement; domain knowledge
-we propose a framework
+Cases such as COMPAS underscore the critical importance of rigorously examining the fairness of AI systems by third parties.
+Fairness is fundementally a subjective social construct, heavily influenced by cultural context and deeply rooted in historical inequalities.
+
+We leverage quantitative measures that provide transparent and objective insights.
+By employing mathematically rigorous methodologies, our research can offer a more deeper look into the fairness of AI systems, enabling third parties to make more informed and robust judgment.
+For example, it would reveal that the COMPAS system violates a fairness measure called equalized odds.
+
+In addition to equalized odds, there are a number of fairness measures for AI systems.
+Our framework aims to help third parties understand these metrics and navigate complex datasets with ease, pinpointing instances of bias and discrimination.
+
+With our tool, a third party such as ProPublica could clearly and easily demonstrates that COMPAS violates equalized odds according to their datasets.
+Our platform empowers organizations to conduct thorough assessments of AI fairness and uncover discrepancies in predictive models.
+
+This functionality not only facilitates sound judgment but also fosters objectivity and transparency regarding AI systems, ultimately bolster the claim of fairness and justice in AI-enabled decision-making.
+
+Our tool is written in Python and is offered as a Python package. It supports common datasets format such as csv. It is open-sourced and publicly available for downloads.
 

### 2024-04-25 abstract.tex
Added author

@@
 
-\title{Ensuring Fairness with Transparent Auditing of Ethical Bias in AI Systems\\
+\title{Ensuring Fairness with Transparent Auditing of Quantitative Bias in AI Systems\\
 \thanks{Identify applicable funding agency here. If none, delete this.}
@@
 
-\author{\IEEEauthorblockN{1\textsuperscript{st} Given Name Surname}
-\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-\textit{name of organization (of Aff.)}\\
-City, Country \\
-email address or ORCID}
+\author{\IEEEauthorblockN{1\textsuperscript{st} Chih-cheng Rex Yuan}
+\IEEEauthorblockA{\textit{Institute of Information Science} \\
+\textit{Academia Sinica}\\
+Taipei, Taiwan \\
+hello@rexyuan.com}
 \and
-\IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
-\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-\textit{name of organization (of Aff.)}\\
-City, Country \\
-email address or ORCID}
+\IEEEauthorblockN{2\textsuperscript{nd} Bow-yaw Wang}
+\IEEEauthorblockA{\textit{Institute of Information Science} \\
+\textit{Academia Sinica}\\
+Taipei, Taiwan \\
+bywang@iis.sinica.edu.tw}
 }

### 2024-04-25 abstract.tex
Added author
Word change

@@
 
-\author{\IEEEauthorblockN{1\textsuperscript{st} Chih-cheng Rex Yuan}
+\author{\IEEEauthorblockN{1\textsuperscript{st} Chih-Cheng Rex Yuan}
 \IEEEauthorblockA{\textit{Institute of Information Science} \\
@@
 \and
-\IEEEauthorblockN{2\textsuperscript{nd} Bow-yaw Wang}
+\IEEEauthorblockN{2\textsuperscript{nd} Bow-Yaw Wang}
 \IEEEauthorblockA{\textit{Institute of Information Science} \\
@@
 In addition to equalized odds, there are a number of fairness measures for AI systems.
-Our framework aims to help third parties understand these metrics and navigate complex datasets with ease, pinpointing instances of bias and discrimination.
+Our framework aims to help third parties apply these metrics and navigate complex datasets with ease, pinpointing instances of bias and discrimination.
 

### 2024-04-25 abstract.tex
Removed author from pdf

--- 2024-04-25 abstract.tex
@@
 }
+
+% \author{\IEEEauthorblockN{1\textsuperscript{st} Chih-Cheng Rex Yuan}
+% \IEEEauthorblockA{\textit{Institute of Information Science} \\
+% \textit{Academia Sinica}\\
+% Taipei, Taiwan \\
+% hello@rexyuan.com}
+% \and
+% \IEEEauthorblockN{2\textsuperscript{nd} Bow-Yaw Wang}
+% \IEEEauthorblockA{\textit{Institute of Information Science} \\
+% \textit{Academia Sinica}\\
+% Taipei, Taiwan \\
+% bywang@iis.sinica.edu.tw}
+% }
 

### 2024-04-26 abstract.tex
Slight change

@@
 
+\begin{IEEEkeywords}
+AI, fairness, auditing
+\end{IEEEkeywords}
+
 \section{Introduction}
@@
 In healthcare, AI-driven diagnostic systems streamline disease identification, while in finance, automated trading algorithms analyze market trends to execute optimal trades swiftly.
-This remarkable progress has reached diverse industries,sprouting out various applications for different purposes.
+This remarkable progress has reached diverse industries, sprouting out various applications for different purposes.
 
@@
 Fairness is fundementally a subjective social construct, heavily influenced by cultural context and deeply rooted in historical inequalities.
+However, there have been development in research leveraging statistical metrics to quantify fairness that provide transparent and objective insights.
 
-We leverage quantitative measures that provide transparent and objective insights.
-By employing mathematically rigorous methodologies, our research can offer a more deeper look into the fairness of AI systems, enabling third parties to make more informed and robust judgment.
+By employing such mathematically rigorous methodologies, our research can offer a more deeper look into the fairness of AI systems, enabling third parties to make more informed and robust judgment.
 For example, it would reveal that the COMPAS system violates a fairness measure called equalized odds.

### 2024-05-19 abstract.tex
Added slides

@@
 \IEEEoverridecommandlockouts
+
+\usepackage[utf8]{inputenc}
+\usepackage[T1]{fontenc}
+
+\usepackage{CJKutf8}
+\usepackage{datetime}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{mathtools}
+\usepackage{subcaption}
+\usepackage{xcolor}
+\usepackage{ulem}
+\usepackage{hyperref}
+\usepackage{graphicx}
 
@@
 }
+
+\DeclarePairedDelimiter{\set}{\{}{\}}
+\DeclarePairedDelimiter{\tuple}{(}{)}
+\DeclarePairedDelimiter{\abs}{\lvert}{\rvert}
+
+\renewcommand{\implies}{\rightarrow}
+
+\newcommand{\bye}[1]{}
 
@@
 
+\begin{table}[h]
+    \centering
+    % \resizebox{\textwidth}{!}{
+    \resizebox{\columnwidth}{!}{
+    \begin{tabular}{|l|c|} % 'l' specifies left alignment for the first column
+        \hline
+        \textbf{Column 1} & \textbf{Column 2} \\
+        \hline
+disparate impact & $\frac{P[\hat{Y} = 1 | S \neq 1]}{P[\hat{Y} = 1 | S = 1]} \geq 1 - \epsilon$ \\
+\hline
+demographic parity & $\abs{P[\hat{Y} = 1 | S = 1] - P[\hat{Y} = 1 | S \neq 1]} \leq \epsilon$ \\
+\hline
+equalized odds: false positive & $\abs{P[\hat{Y} = 1 | S = 1, Y = 0] - P[\hat{Y} = 1 | S \neq 1, Y = 0]} \leq \epsilon$ \\
+\hline
+equalized odds: true positive & $\abs{P[\hat{Y} = 1 | S = 1, Y = 1] - P[\hat{Y} = 1 | S \neq 1, Y = 1]} \leq \epsilon$ \\
+\hline
+equal opportunity & $\abs{P[\hat{Y} = 1 | S = 1, Y = 1] - P[\hat{Y} = 1 | S \neq 1, Y = 1]} \leq \epsilon$ \\
+\hline
+accuracy eqaulity & $\abs{P[Y = \hat{Y} | S = 1] - P[Y = \hat{Y} | S \neq 1]} \leq \epsilon$ \\
+\hline
+predictive parity & $\abs{P[Y = 1 | S = 1, \hat{Y} = 1] - P[Y = 1 | S \neq 1, \hat{Y} = 1]} \leq \epsilon$ \\
+\hline
+equal calibration & $\abs{P[Y = 1 | S = 1, \hat{V} = v] - P[Y = 1 | S \neq 1, \hat{V} = v]} \leq \epsilon$ \\
+\hline
+conditional statistical parity & $\abs{P[\hat{Y} = 1 | S = 1, L = l] - P[\hat{Y} = 1 | S \neq 1, L = l]} \leq \epsilon$ \\
+\hline
+predictive equality & $\abs{P[\hat{Y} = 1 | S = 1, Y = 0] - P[\hat{Y} = 1 | S \neq 1, Y = 0]} \leq \epsilon$ \\
+\hline
+conditional use accuracy equaliyt: true positive & $\abs{P[Y = 1 | S = 1, \hat{Y} = 1] - P[Y = 1 | S \neq 1, \hat{Y} = 1]} \leq \epsilon$ \\
+\hline
+conditional use accuracy equality: true negative & $\abs{P[Y = 0 | S = 1, \hat{Y} = 0] - P[Y = 0 | S \neq 1, \hat{Y} = 0]} \leq \epsilon$ \\
+\hline
+mean difference & $\abs{E[\hat{Y}|S = 1] - E[\hat{Y}|S \neq 1]} \leq \epsilon$ \\
+        \hline
+    \end{tabular}
+    }
+    \caption{Example Table}
+    \label{tab:example}
+\end{table}
+
 \end{document}

### 2024-05-21 abstract.tex
Added reference

@@
 \usepackage{graphicx}
+\usepackage{cite}
 
@@
 
-\begin{table}[h]
+\section{Fairness Measures}
+Fairness is about making sure the disadvantaged and unprivileged groups of individuals are treated equitably.
+However, there have been several conceptions of it proposed in the past.
+To have a constructive discussion on fairness, we must first have precise definitions of it.
+Pessach and Shmueli \cite{pessach2022review} formulated a number of fairness measures in an unified mathematical notation.
+We will base our framework on their formulation.
+
+\subsection{Preliminaries}
+For the following definition we will use $Y$ to denote the ground truth of an outcome; $\hat{Y}$ to denote the predicated result of an outcome; $Y = 1$ and $\hat{Y} = 1$ to denote them being accepted or positive.
+For example, let $Y$ be recidivism.
+Then $Y = 1$ means the case of an individual actually recidivating and $\hat{Y} = 0$ means the predication of an individual recidivating is negative.
+
+We denote by $S$ some protected attribute.
+Protected attributes are the characteristics of individuals that are, for example, legally or ethically, considered sensitive and warrant protection against discrimination and bias.
+We write $S = 1$ to represent the privileged group and $S \neq 1$ to represent the unprivileged group.
+For example, let $S$ be Caucasian.
+Then $S = 1$ represents the case of an individual's race being Caucasian and $S \neq 1$ vice versa.
+
+\subsection{Disparate Impact}
+In 1971 \cite{griggs1971}, the US spreme court ruled that it is illegal for hiring decisions to have  ``disparate impact'' by race, thus coining the term.
+It is taken as unintentional discrimination, as opposed to intentional discrimination, which is called ``disparate treatment''.
+
+Legal cases involving disparate impact often refer to the ``80\% Rule'', advocated by the US Equal Employment Opportunity Commission \cite{eeoc1979}, where it requires the selection rate of a minority group is to be no less than 80\% of that of a majority group.
+Formally, it is:
+
+\[
+    \frac{P[\hat{Y} = 1 | S \neq 1]}{P[\hat{Y} = 1 | S = 1]} \geq 1 - \epsilon
+\]
+
+\subsection{Demographic Parity}
+Demographic parity, also known as statistical parity, is similar to disparate impact but, instead of ratio, difference is taken.
+It is named so to suggest that each demographic group(such as race, gender, or age) should have equal representation or opportunity.
+Formally, it is:
+
+\[
+    \abs{P[\hat{Y} = 1 | S = 1] - P[\hat{Y} = 1 | S \neq 1]} \leq \epsilon
+\]
+
+\begin{table*}[h]
     \centering
     % \resizebox{\textwidth}{!}{
-    \resizebox{\columnwidth}{!}{
+    % \resizebox{\columnwidth}{!}{
     \begin{tabular}{|l|c|} % 'l' specifies left alignment for the first column
@@
     \end{tabular}
-    }
+    % }
     \caption{Example Table}
     \label{tab:example}
-\end{table}
+\end{table*}
+
+\bye{
+fairness definition 1p
+    example
+
+framework definition 1p
+    abstraction
+
+framework example usage 1p
+    csv input
+    model input
+
+compas introduction
+
+compas analysis
+}
+
+\bibliographystyle{IEEEtran}
+\bibliography{references}
 

### 2024-05-21 abstract.tex
Added some more defs

@@
 \usepackage{cite}
+\usepackage{multirow}
 
@@
 Fairness is about making sure the disadvantaged and unprivileged groups of individuals are treated equitably.
-However, there have been several conceptions of it proposed in the past.
+However, there have been several interpretations of it proposed in the past.
 To have a constructive discussion on fairness, we must first have precise definitions of it.
@@
 Then $Y = 1$ means the case of an individual actually recidivating and $\hat{Y} = 0$ means the predication of an individual recidivating is negative.
+In addition, we use $V$ and $\hat{V}$ when the truth and the prediction isn't binary.
+For example, we denote COMPAS score by $\hat{V}$, which ranges from 1 to 10.
 
@@
 
+\subsection{Conditional Statistical Parity}
+Conditional statistical parity is similar to demographic parity, but, in addition to protected attributes, it further takes into account some ``legitimate'' attributes that are \textit{legitimately} related to the case.
+For example, a legitimate attribute when considering future recidivism could be the number of prior crimes committed.
+Formally, it is:
+
+\[
+    \abs{P[\hat{Y} = 1 | S = 1, L = l] - P[\hat{Y} = 1 | S \neq 1, L = l]} \leq \epsilon
+\]
+
+where $L$ to denotes the legitimate attributes.
+
+\subsection{Overall Accuracy Equality}
+Accuracy eqaulity is similar to demographic parity, but instead of the case of $\hat{Y} = 1$, it considers the case of $Y = \hat{Y}$; that is, the case where the prediction is accurate.
+Formally, it is:
+
+\[
+    \abs{P[Y = \hat{Y} | S = 1] - P[Y = \hat{Y} | S \neq 1]} \leq \epsilon
+\]
+
+\subsection{Mean Difference}
+Mean difference considers the case of non-binary outcome, such as the score of COMPAS. It takes the expected value of the prediction. Formally, it is:
+
+\[
+    \abs{E[\hat{Y}|S = 1] - E[\hat{Y}|S \neq 1]} \leq \epsilon
+\]
+
 \begin{table*}[h]
@@
 \hline
-equalized odds: false positive & $\abs{P[\hat{Y} = 1 | S = 1, Y = 0] - P[\hat{Y} = 1 | S \neq 1, Y = 0]} \leq \epsilon$ \\
+conditional statistical parity & $\abs{P[\hat{Y} = 1 | S = 1, L = l] - P[\hat{Y} = 1 | S \neq 1, L = l]} \leq \epsilon$ \\
 \hline
-equalized odds: true positive & $\abs{P[\hat{Y} = 1 | S = 1, Y = 1] - P[\hat{Y} = 1 | S \neq 1, Y = 1]} \leq \epsilon$ \\
+accuracy eqaulity & $\abs{P[Y = \hat{Y} | S = 1] - P[Y = \hat{Y} | S \neq 1]} \leq \epsilon$ \\
+\hline
+mean difference & $\abs{E[\hat{Y}|S = 1] - E[\hat{Y}|S \neq 1]} \leq \epsilon$ \\
+\hline
+\multirow{2}{*}{equalized odds} & $\abs{P[\hat{Y} = 1 | S = 1, Y = 0] - P[\hat{Y} = 1 | S \neq 1, Y = 0]} \leq \epsilon$ \\
+\cline{2-2}
+                                & $\abs{P[\hat{Y} = 1 | S = 1, Y = 1] - P[\hat{Y} = 1 | S \neq 1, Y = 1]} \leq \epsilon$ \\
 \hline
@@
 \hline
-accuracy eqaulity & $\abs{P[Y = \hat{Y} | S = 1] - P[Y = \hat{Y} | S \neq 1]} \leq \epsilon$ \\
+predictive equality & $\abs{P[\hat{Y} = 1 | S = 1, Y = 0] - P[\hat{Y} = 1 | S \neq 1, Y = 0]} \leq \epsilon$ \\
+\hline
+\multirow{2}{*}{conditional use accuracy equality} & $\abs{P[Y = 1 | S = 1, \hat{Y} = 1] - P[Y = 1 | S \neq 1, \hat{Y} = 1]} \leq \epsilon$ \\
+\cline{2-2}
+                                 & $\abs{P[Y = 0 | S = 1, \hat{Y} = 0] - P[Y = 0 | S \neq 1, \hat{Y} = 0]} \leq \epsilon$ \\
 \hline
@@
 equal calibration & $\abs{P[Y = 1 | S = 1, \hat{V} = v] - P[Y = 1 | S \neq 1, \hat{V} = v]} \leq \epsilon$ \\
-\hline
-conditional statistical parity & $\abs{P[\hat{Y} = 1 | S = 1, L = l] - P[\hat{Y} = 1 | S \neq 1, L = l]} \leq \epsilon$ \\
-\hline
-predictive equality & $\abs{P[\hat{Y} = 1 | S = 1, Y = 0] - P[\hat{Y} = 1 | S \neq 1, Y = 0]} \leq \epsilon$ \\
-\hline
-conditional use accuracy equaliyt: true positive & $\abs{P[Y = 1 | S = 1, \hat{Y} = 1] - P[Y = 1 | S \neq 1, \hat{Y} = 1]} \leq \epsilon$ \\
-\hline
-conditional use accuracy equality: true negative & $\abs{P[Y = 0 | S = 1, \hat{Y} = 0] - P[Y = 0 | S \neq 1, \hat{Y} = 0]} \leq \epsilon$ \\
-\hline
-mean difference & $\abs{E[\hat{Y}|S = 1] - E[\hat{Y}|S \neq 1]} \leq \epsilon$ \\
 \hline

### 2024-05-21 abstract.tex
Added some more defs

@@
 
+We denote by $\epsilon$ some threshold that's used to limit the fairness measures.
+
 \subsection{Disparate Impact}
@@
 \]
+
+In the case of 80\% rule, $\epsilon = 20\%$.
 
@@
 
-where $L$ to denotes the legitimate attributes.
+where $L$ denotes the legitimate attributes.
 
@@
 \]
+
+\subsection{Equalized Odds}
+Equalized odds is similar to demogrphic parity, but it further takes into account the grouth truth.
+It consider the cases of true positive and false positive.
+It solves the downsides of a fully accurate classifier might be deemed unfair by measures that don't take ground truth into consideration.
+For example, consider a group $A$ that is predicated to recidivate and they do in fact always recidivate and a group $B$ that is predicated to recidivate but end up never recidivating.
+A fully accurate classifier will always predict $A$ to recidivate while $B$ to never recidivate, and this will violate demogrphic parity.
+By taking ground truth into account, equalized odds avoids these pitfalls.
+Formally, it is:
+
+\begin{align*}
+    \abs{P[\hat{Y} = 1 | S = 1, Y = 0] - P[\hat{Y} = 1 | S \neq 1, Y = 0]} & \leq \epsilon \\
+    \abs{P[\hat{Y} = 1 | S = 1, Y = 1] - P[\hat{Y} = 1 | S \neq 1, Y = 1]} & \leq \epsilon
+\end{align*}
+
+\subsection{Equal Opportunity}
+Equal opportunity is a relaxation of equalized odds by only considering the true positive case.
+It is named so because when outcome being positive is beneficial to the individual, as it is an opportunity, such as school acceptance, the fairness of the true positive case is much more important.
+Formally, it is:
+
+\begin{align*}
+    \abs{P[\hat{Y} = 1 | S = 1, Y = 1] - P[\hat{Y} = 1 | S \neq 1, Y = 1]} & \leq \epsilon
+\end{align*}
+
+\subsection{Predictive Equality}
+Predictive equality is also a relaxation of equalized odds by only considering the false positive case.
+It was formulated \cite{corbett2017algorithmic} when considering outcome being positive is detrimental to the individual, such as future recidivism.
+Thus, the fairness of the false positive case is much more important.
+Formally, it is:
+
+\begin{align*}
+    \abs{P[\hat{Y} = 1 | S = 1, Y = 0] - P[\hat{Y} = 1 | S \neq 1, Y = 0]} \leq \epsilon
+\end{align*}
 

### 2024-05-22 abstract.tex
Added framework

@@
 \renewcommand{\implies}{\rightarrow}
+
+\newcommand{\db}{\mathcal{D}}
+\newcommand{\model}{\mathcal{M}}
+\newcommand{\priv}{P}
+\newcommand{\pos}{\hat{P}}
+\newcommand{\tru}{T}
+\newcommand{\leg}{L}
+\newcommand{\sco}{\hat{S}}
 
@@
 Legal cases involving disparate impact often refer to the ``80\% Rule'', advocated by the US Equal Employment Opportunity Commission \cite{eeoc1979}, where it requires the selection rate of a minority group is to be no less than 80\% of that of a majority group.
-Formally, it is:
+Formally \cite{feldman2015certifying}, it is:
 
@@
 
-\subsection{Demographic Parity}
-Demographic parity, also known as statistical parity, is similar to disparate impact but, instead of ratio, difference is taken.
+\subsection{Demographic Parity} %TODO: why cite this?
+Demographic parity \cite{calders2010three}, also known as statistical parity, is similar to disparate impact but, instead of ratio, difference is taken.
 It is named so to suggest that each demographic group(such as race, gender, or age) should have equal representation or opportunity.
@@
 \subsection{Conditional Statistical Parity}
-Conditional statistical parity is similar to demographic parity, but, in addition to protected attributes, it further takes into account some ``legitimate'' attributes that are \textit{legitimately} related to the case.
+Conditional statistical parity \cite{corbett2017algorithmic} is similar to demographic parity, but, in addition to protected attributes, it further takes into account some ``legitimate'' attributes that are \textit{legitimately} related to the case.
 For example, a legitimate attribute when considering future recidivism could be the number of prior crimes committed.
@@
 \subsection{Overall Accuracy Equality}
-Accuracy eqaulity is similar to demographic parity, but instead of the case of $\hat{Y} = 1$, it considers the case of $Y = \hat{Y}$; that is, the case where the prediction is accurate.
+Overall accuracy eqaulity \cite{berk2021fairness} is similar to demographic parity, but instead of the case of $\hat{Y} = 1$, it considers the case of $Y = \hat{Y}$; that is, the case where the prediction is accurate.
 Formally, it is:
@@
 \subsection{Mean Difference}
-Mean difference considers the case of non-binary outcome, such as the score of COMPAS. It takes the expected value of the prediction. Formally, it is:
+Mean difference \cite{vzliobaite2017measuring} considers the case of non-binary outcome, such as the score of COMPAS. It takes the expected value of the prediction. Formally, it is:
 
@@
 \subsection{Equalized Odds}
-Equalized odds is similar to demogrphic parity, but it further takes into account the grouth truth.
+Equalized odds \cite{hardt2016equality} is similar to demogrphic parity, but it further takes into account the grouth truth.
 It consider the cases of true positive and false positive.
@@
 \subsection{Equal Opportunity}
-Equal opportunity is a relaxation of equalized odds by only considering the true positive case.
+Equal opportunity \cite{hardt2016equality} is a relaxation of equalized odds by only considering the true positive case.
 It is named so because when outcome being positive is beneficial to the individual, as it is an opportunity, such as school acceptance, the fairness of the true positive case is much more important.
@@
 \subsection{Predictive Equality}
-Predictive equality is also a relaxation of equalized odds by only considering the false positive case.
-It was formulated \cite{corbett2017algorithmic} when considering outcome being positive is detrimental to the individual, such as future recidivism.
-Thus, the fairness of the false positive case is much more important.
+Predictive equality \cite{corbett2017algorithmic} is also a relaxation of equalized odds by only considering the false positive case.
+It was formulated by considering positive outcome being detrimental to the individual, such as predicting future recidivism.
+Thus, only the false positive case is taken.
 Formally, it is:
@@
 \end{align*}
+
+\subsection{Conditional Use Accuracy Equality}
+Conditional use accuracy equality \cite{berk2021fairness} is similar to equalized odds, but instead of conditioning on the ground truth, it conditions on the prediction and calculated the probability of the ground truth.
+It can be seem as checking the prediction accuracy across groups, thus the name.
+It further requires the measure on the case of positive predictive values is less than that of the negative predictive values.
+Formally, it is:
+
+\begin{align*}
+    \abs{P[Y = 1 | S = 1, \hat{Y} = 1] & - P[Y = 1 | S \neq 1, \hat{Y} = 1]} && \leq \epsilon \\
+    & \rotatebox{90}{>} && \\
+    \abs{P[Y = 0 | S = 1, \hat{Y} = 0] & - P[Y = 0 | S \neq 1, \hat{Y} = 0]} && \leq \epsilon
+\end{align*}
+
+\subsection{Predictive Parity}
+Predictive parity \cite{chouldechova2017fair} is a relaxation of conditional use accuracy equality by only considering the positive predictive value case.
+Like predictive equality, it was formulated by considering positive outcome being detrimental to the individual.
+Thus, only the positive predictive value case is taken.
+Formally, it is:
+
+\begin{align*}
+    \abs{P[Y = 1 | S = 1, \hat{Y} = 1] - P[Y = 1 | S \neq 1, \hat{Y} = 1]} \leq \epsilon
+\end{align*}
+
+\subsection{Equal Calibration}
+Equal calibration \cite{chouldechova2017fair} is similar to equal opportunity, but instead of having a binary $\hat{Y}$, it is conditioned on the range of the predicted value $\hat{V}$.
+For example, this could be conditioned on the highest COMPAS score $\hat{V} = 10$.
+Calibration is a concept of having fair score function \cite{fraenkel2020fairness}.
+Formally, it is:
+
+\begin{align*}
+    \abs{P[Y = 1 | S = 1, \hat{V} = v] - P[Y = 1 | S \neq 1, \hat{V} = v]} \leq \epsilon
+\end{align*}
+
+\subsection{Positive Balance}
+Positive balance \cite{kleinberg2016inherent} is similar to equal opportunity, but instead of taking the difference of probability of binary prediction $\hat{Y}$, it takes the difference of the expected value of the score $\hat{V}$.
+Formally, it is:
+
+\begin{align*}
+    \abs{E[\hat{V} | Y = 1, S = 1] - E[\hat{V} | Y = 1, S \neq 1]} \leq \epsilon
+\end{align*}
+
+\subsection{Negative Balance}
+Negative balance \cite{kleinberg2016inherent} is like positive balance except it conditions on the case of $Y = 0$. Formally, it is:
+
+\begin{align*}
+    \abs{E[\hat{V} | Y = 0, S = 1] - E[\hat{V} | Y = 0, S \neq 1]} \leq \epsilon
+\end{align*}
+
+These fairness measures are compiled in Table~\ref{tab:measures}.
 
@@
         \hline
-disparate impact & $\frac{P[\hat{Y} = 1 | S \neq 1]}{P[\hat{Y} = 1 | S = 1]} \geq 1 - \epsilon$ \\
+Disparate Impact & $\frac{P[\hat{Y} = 1 | S \neq 1]}{P[\hat{Y} = 1 | S = 1]} \geq 1 - \epsilon$ \\
 \hline
-demographic parity & $\abs{P[\hat{Y} = 1 | S = 1] - P[\hat{Y} = 1 | S \neq 1]} \leq \epsilon$ \\
+Demographic Parity & $\abs{P[\hat{Y} = 1 | S = 1] - P[\hat{Y} = 1 | S \neq 1]} \leq \epsilon$ \\
 \hline
-conditional statistical parity & $\abs{P[\hat{Y} = 1 | S = 1, L = l] - P[\hat{Y} = 1 | S \neq 1, L = l]} \leq \epsilon$ \\
+Conditional Statistical Parity & $\abs{P[\hat{Y} = 1 | S = 1, L = l] - P[\hat{Y} = 1 | S \neq 1, L = l]} \leq \epsilon$ \\
 \hline
-accuracy eqaulity & $\abs{P[Y = \hat{Y} | S = 1] - P[Y = \hat{Y} | S \neq 1]} \leq \epsilon$ \\
+Overall Accuracy Eqaulity & $\abs{P[Y = \hat{Y} | S = 1] - P[Y = \hat{Y} | S \neq 1]} \leq \epsilon$ \\
 \hline
-mean difference & $\abs{E[\hat{Y}|S = 1] - E[\hat{Y}|S \neq 1]} \leq \epsilon$ \\
+Mean Difference & $\abs{E[\hat{Y}|S = 1] - E[\hat{Y}|S \neq 1]} \leq \epsilon$ \\
 \hline
-\multirow{2}{*}{equalized odds} & $\abs{P[\hat{Y} = 1 | S = 1, Y = 0] - P[\hat{Y} = 1 | S \neq 1, Y = 0]} \leq \epsilon$ \\
+\multirow{2}{*}{Equalized Odds} & $\abs{P[\hat{Y} = 1 | S = 1, Y = 0] - P[\hat{Y} = 1 | S \neq 1, Y = 0]} \leq \epsilon$ \\
 \cline{2-2}
@@
 \hline
-equal opportunity & $\abs{P[\hat{Y} = 1 | S = 1, Y = 1] - P[\hat{Y} = 1 | S \neq 1, Y = 1]} \leq \epsilon$ \\
+Equal Opportunity & $\abs{P[\hat{Y} = 1 | S = 1, Y = 1] - P[\hat{Y} = 1 | S \neq 1, Y = 1]} \leq \epsilon$ \\
 \hline
-predictive equality & $\abs{P[\hat{Y} = 1 | S = 1, Y = 0] - P[\hat{Y} = 1 | S \neq 1, Y = 0]} \leq \epsilon$ \\
+Predictive Equality & $\abs{P[\hat{Y} = 1 | S = 1, Y = 0] - P[\hat{Y} = 1 | S \neq 1, Y = 0]} \leq \epsilon$ \\
 \hline
-\multirow{2}{*}{conditional use accuracy equality} & $\abs{P[Y = 1 | S = 1, \hat{Y} = 1] - P[Y = 1 | S \neq 1, \hat{Y} = 1]} \leq \epsilon$ \\
+\multirow{2}{*}{Conditional Use Accuracy Equality} & $\abs{P[Y = 1 | S = 1, \hat{Y} = 1] - P[Y = 1 | S \neq 1, \hat{Y} = 1]} \leq \epsilon$ \\
 \cline{2-2}
@@
 \hline
-predictive parity & $\abs{P[Y = 1 | S = 1, \hat{Y} = 1] - P[Y = 1 | S \neq 1, \hat{Y} = 1]} \leq \epsilon$ \\
+Predictive Parity & $\abs{P[Y = 1 | S = 1, \hat{Y} = 1] - P[Y = 1 | S \neq 1, \hat{Y} = 1]} \leq \epsilon$ \\
 \hline
-equal calibration & $\abs{P[Y = 1 | S = 1, \hat{V} = v] - P[Y = 1 | S \neq 1, \hat{V} = v]} \leq \epsilon$ \\
+Equal Calibration & $\abs{P[Y = 1 | S = 1, \hat{V} = v] - P[Y = 1 | S \neq 1, \hat{V} = v]} \leq \epsilon$ \\
+\hline
+Positive Balance & $\abs{E[\hat{V} | Y = 1, S = 1] - E[\hat{V} | Y = 1, S \neq 1]} \leq \epsilon$ \\
+\hline
+Negative Balance & $\abs{E[\hat{V} | Y = 0, S = 1] - E[\hat{V} | Y = 0, S \neq 1]} \leq \epsilon$ \\
 \hline
@@
     % }
-    \caption{Example Table}
-    \label{tab:example}
+    \caption{Fairness measures.}
+    \label{tab:measures}
 \end{table*}
+
+\section{Auditing Framework}
+Per the review by Pessach and Shmueli \cite{pessach2022review}, we designed an auditing framework for calculating the various fairness measure.
+We offer two versions of fairness checkers: one for when the prediction positive results are readily available in csv input and one for when a model is provided.
+In most auditing cases the model version is preferred because csv result can be easily fabricated.
+
+\subsection{Preliminaries}
+Let $\db = \set{r_1, r_2, ...}$ be a database containing rows of data.
+Each row $r_i : \mathtt{str} \implies \mathtt{str}$ is a lookup table or dictionary.
+For example, $r_n(\text{``race''}) = \text{``Asian''}$ means $r_n$'s race is Asian.
+
+A model $\model : (\mathtt{str} \implies \mathtt{str}) \implies \alpha$ takes a row and returns the model's prediction result of some type $\alpha$.
+
+A privileged predicate $\priv : (\mathtt{str} \implies \mathtt{str}) \implies \mathtt{bool}$ takes a row and determines if it belongs to the privileged group.
+
+A positive predicate of rows $\pos : (\mathtt{str} \implies \mathtt{str}) \implies \mathtt{bool}$ takes a row and determines if its prediction is positive.
+A positive predicate of model results $\pos : \alpha \implies \mathtt{bool}$ takes a model's result and determines if it is positive.
+
+A ground truth predicate $\tru : (\mathtt{str} \implies \mathtt{str}) \implies \mathtt{bool}$ takes a row and gives the ground truth of the result.
+
+A legitimate predicate $\leg : \alpha_i^n \implies (\mathtt{str} \implies \mathtt{str}) \implies \mathtt{bool}$, and similarly a score predicate $\sco : \alpha_i^n \implies (\mathtt{str} \implies \mathtt{str}) \implies \mathtt{bool}$, takes $n$ parameters and returns a row predicate.
+
+\subsection{Definitions}
+We abstracted the idea of privileged groups and positive prediction as predicates to maximizes flexibility.
+With the proper predicates, any model output can be used; even prose like response of generative models can be included given adequate predicates.
+
+Given a datasets for auditing use, and given csv prediction results or the prediction model itself, we can calculate the fairness measures by:
+
+\begin{align*}
+\text{disparate\_impact} & (\epsilon, \model, \priv, \pos) \\
+\text{demographic\_parity} & (\epsilon, \model, \priv, \pos) \\
+\text{conditional\_statistical\_parity} & (\epsilon, \model, \priv, \pos, \leg, (args, ...)) \\
+\text{overall\_accuracy\_eqaulity} & (\epsilon, \model, \priv, \pos, \tru) \\
+\text{mean\_difference} & (\epsilon, \model, \priv, \pos) \\
+\text{equalized\_odds} & (\epsilon, \model, \priv, \pos, \tru) \\
+\text{equal\_opportunity} & (\epsilon, \model, \priv, \pos, \tru) \\
+\text{predictive\_equality} & (\epsilon, \model, \priv, \pos, \tru) \\
+\text{conditional\_use\_accuracy\_equality} & (\epsilon, \model, \priv, \pos, \tru) \\
+\text{predictive\_parity} & (\epsilon, \model, \priv, \pos, \tru) \\
+\text{equal\_calibration} & (\epsilon, \model, \priv, \sco, \tru, (args, ...)) \\
+\text{positive\_balance} & (\epsilon, \model, \priv, \sco, \tru, (args, ...)) \\
+\text{nagative\_balance} & (\epsilon, \model, \priv, \sco, \tru, (args, ...)) \\
+\end{align*}
+
+where $\mathcal{M}$ is optional if the input contains csv prediction results.
 

### 2024-05-23 paper/abstract.tex
Reorganized files

--- 2024-05-22 abstract.tex
@@
 \section{Fairness Measures}
+\label{sec:measures}
 Fairness is about making sure the disadvantaged and unprivileged groups of individuals are treated equitably.
@@
         \hline
-        \textbf{Column 1} & \textbf{Column 2} \\
+        \textbf{Fairness Measure} & \textbf{Definition} \\
         \hline
@@
 \section{Auditing Framework}
+\label{sec:framework}
 Per the review by Pessach and Shmueli \cite{pessach2022review}, we designed an auditing framework for calculating the various fairness measure.
@@
 Each row $r_i : \mathtt{str} \implies \mathtt{str}$ is a lookup table or dictionary.
-For example, $r_n(\text{``race''}) = \text{``Asian''}$ means $r_n$'s race is Asian.
+For example, $r_n(\text{``sex''}) = \text{``Female''}$ means $r_n$'s sex is female.
 
@@
 A privileged predicate $\priv : (\mathtt{str} \implies \mathtt{str}) \implies \mathtt{bool}$ takes a row and determines if it belongs to the privileged group.
+For example, $\priv(r_i) := r_i(\text{``race''}) == \text{``Caucasian''}$ means the privileged group is those with race being Caucasian.
 
 A positive predicate of rows $\pos : (\mathtt{str} \implies \mathtt{str}) \implies \mathtt{bool}$ takes a row and determines if its prediction is positive.
+For example, $\pos(r_i) := \mathtt{int}(r_i(\text{``score''})) > \text{``7''}$ means a row's prediction is positive if its score is greater than 7.
 A positive predicate of model results $\pos : \alpha \implies \mathtt{bool}$ takes a model's result and determines if it is positive.
+In the simplest case, the model returns a $\mathtt{bool}$, and the positive predicate can just be the idnetity function.
 
@@
 A legitimate predicate $\leg : \alpha_i^n \implies (\mathtt{str} \implies \mathtt{str}) \implies \mathtt{bool}$, and similarly a score predicate $\sco : \alpha_i^n \implies (\mathtt{str} \implies \mathtt{str}) \implies \mathtt{bool}$, takes $n$ parameters and returns a row predicate.
+For example, $\sco(u, l)(r_i) := l < \mathtt{int}(r_i(\text{``score''})) < u$ first takes two parameters $u,l$ as upper bound and lower bound, and then decides if $r_i$'s prediction score is between them.
 
@@
 We abstracted the idea of privileged groups and positive prediction as predicates to maximizes flexibility.
-With the proper predicates, any model output can be used; even prose like response of generative models can be included given adequate predicates.
+With the proper predicates, any model output can be used; even prose-like response of generative models can be included given adequate predicates.
 
-Given a datasets for auditing use, and given csv prediction results or the prediction model itself, we can calculate the fairness measures by:
+Given a datasets for auditing use, and given csv prediction results or the prediction model itself, we can calculate the fairness measures mentioned in Section~\ref{sec:measures} by:
 
@@
 where $\mathcal{M}$ is optional if the input contains csv prediction results.
+
+By calculating all of the available fairness measures, we provide the auditors a more comprehensive perespective on a model or dataset.
+
+\section{Application}
+\label{sec:app}
+In this section, we will apply the framework proposed in Section~\ref{sec:framework} to the ProPublica COMPAS dataset.
+
+\subsection{ProPublica COMPAS Dataset}
+Correctional Offender Management Profiling for Alternative Sanctions (COMPAS), developed by private company Northpointe (now Equivant), is a risk assessment software used in the American criminal justice system to evaluate the likelihood of a defendant reoffending.
+
+ProPublica is an American non-profit organization focused on investigative journalism in public interests. In 2016, they conduct a
 

### 2024-05-23 setup.py
Added source code

@@ -1,0 +1,8 @@
+from setuptools import setup, find_packages
+
+setup(
+    name="fairness_checker",
+    version="0.0.1",
+    package_dir={'': 'src'},
+    packages=find_packages(where="src")
+)

### 2024-05-23 test/test_csv.py

@@ -1,0 +1,22 @@
+from fairness_checker import *
+
+race = 'Native American'
+sex = 'Male'
+degree = 'M'
+age = 'Greater than 45'
+
+c = fairness_csv_checker("compas-scores-two-years.csv")
+
+c.disparate_impact(0.8, lambda row:                  row['sex'] == sex, lambda row: row['score_text'] in {'Medium', 'High'})
+c.demographic_parity(0.2, lambda row:                row['sex'] == sex, lambda row: row['score_text'] in {'Medium', 'High'})
+c.equalized_odds(0.2, lambda row:                    row['sex'] == sex, lambda row: row['score_text'] in {'Medium', 'High'}, lambda row: row['is_recid'] == '1')
+c.equal_opportunity(0.2, lambda row:                 row['sex'] == sex, lambda row: row['score_text'] in {'Medium', 'High'}, lambda row: row['is_recid'] == '1')
+c.accuracy_eqaulity(0.2, lambda row:                 row['sex'] == sex, lambda row: row['score_text'] in {'Medium', 'High'}, lambda row: row['is_recid'] == '1')
+c.predictive_parity(0.2, lambda row:                 row['sex'] == sex, lambda row: row['score_text'] in {'Medium', 'High'}, lambda row: row['is_recid'] == '1')
+c.equal_calibration(0.2, lambda row:                 row['sex'] == sex, lambda u,l: (lambda row: l <= int(row['decile_score']) and int(row['decile_score']) <= u), (7, 5), lambda row: row['is_recid'] == '1')
+c.conditional_statistical_parity(0.2, lambda row:    row['sex'] == sex, lambda row: row['score_text'] in {'Medium', 'High'}, lambda x: (lambda row: int(row['priors_count']) > x), (0,))
+c.predictive_equality(0.2, lambda row:               row['sex'] == sex, lambda row: row['score_text'] in {'Medium', 'High'}, lambda row: row['is_recid'] == '1')
+c.conditional_use_accuracy_equality(0.2, lambda row: row['sex'] == sex, lambda row: row['score_text'] in {'Medium', 'High'}, lambda row: row['is_recid'] == '1')
+c.positive_balance(5, lambda row:                    row['sex'] == sex, lambda row: int(row['decile_score']), lambda row: row['is_recid'] == '1')
+c.nagative_balance(5, lambda row:                    row['sex'] == sex, lambda row: int(row['decile_score']), lambda row: row['is_recid'] == '1')
[…]

### 2024-05-23 test/test_model.py

@@ -1,0 +1,128 @@
+from fairness_checker import *
+
+import pandas as pd
+
+from sklearn import preprocessing
+from sklearn.model_selection import train_test_split
+from sklearn.linear_model import LogisticRegression
+from sklearn.tree import DecisionTreeClassifier
+from sklearn.ensemble import RandomForestClassifier
+from sklearn.metrics import accuracy_score
+
+df = pd.read_csv("compas_train.csv")
+df = df.loc[:, ['sex', 'age_cat', 'race', 'priors_count', 'c_charge_degree', 'two_year_recid', 'decile_score']]
+
+sex_encoder = preprocessing.LabelEncoder()
+age_encoder = preprocessing.LabelEncoder()
+race_encoder = preprocessing.LabelEncoder()
+degree_encoder = preprocessing.LabelEncoder()
+
+sex_cat = sex_encoder.fit_transform(df['sex'])
+age_cat = age_encoder.fit_transform(df['age_cat'])
+race_cat = race_encoder.fit_transform(df['race'])
+degree_cat = degree_encoder.fit_transform(df['c_charge_degree'])
+
+sex_lookup = dict(zip(sex_encoder.classes_, sex_encoder.transform(sex_encoder.classes_)))
+age_lookup = dict(zip(age_encoder.classes_, age_encoder.transform(age_encoder.classes_)))
+race_lookup = dict(zip(race_encoder.classes_, race_encoder.transform(race_encoder.classes_)))
+degree_lookup = dict(zip(degree_encoder.classes_, degree_encoder.transform(degree_encoder.classes_)))
+
+df['sex_cat'] = sex_cat
+df['age_cat'] = age_cat
+df['race_cat'] = race_cat
+df['degree_cat'] = degree_cat
+
+dummy_fields = ['sex', 'age_cat', 'race', 'c_charge_degree']
+data = df.drop(dummy_fields, axis = 1)
+
+data = df.reindex(['sex_cat', 'age_cat', 'race_cat', 'degree_cat', 'two_year_recid'], axis=1)
+X_train = data.iloc[:, 0:4]
+Y_train = data.iloc[:, 4]
+
+model = LogisticRegression()
+model.fit(X_train, Y_train)
+
+data = df.reindex(['sex_cat', 'age_cat', 'race_cat', 'degree_cat', 'decile_score'], axis=1)
+X_train = data.iloc[:, 0:4]
+Y_train = data.iloc[:, 4]
+
[…]

### 2024-05-23 setup.py
Added license

@@ -1,8 +1,18 @@
 from setuptools import setup, find_packages
 
+with open("src/README.md", "r") as f:
+    long_description = f.read()
+
 setup(
     name="fairness_checker",
     version="0.0.1",
     package_dir={'': 'src'},
-    packages=find_packages(where="src")
+    packages=find_packages(where="src"),
+    description="Fairnes checker",
+    long_description=long_description,
+    long_description_content_type="text/markdown",
+    url="https://github.com/RexYuan/Shu",
+    author="RexYuan",
+    author_email="hello@rexyuan.com",
+    license="Unlicense",
 )

### 2024-05-24 paper/abstract.tex
Added a bit of analysis

@@
 \usepackage{multirow}
+\usepackage{makecell}
 
@@
 
-By calculating all of the available fairness measures, we provide the auditors a more comprehensive perespective on a model or dataset.
+By calculating all of the available fairness measures, we provide the auditors a comprehensive perespective on the fairness performance of a model or dataset.
 
-\section{Application}
+\subsection{Application}
 \label{sec:app}
-In this section, we will apply the framework proposed in Section~\ref{sec:framework} to the ProPublica COMPAS dataset.
+In this section, we will apply the proposed framework to the ProPublica COMPAS dataset \cite{angwin2016machine,larson2016compas,larson2016propublica}.
 
-\subsection{ProPublica COMPAS Dataset}
 Correctional Offender Management Profiling for Alternative Sanctions (COMPAS), developed by private company Northpointe (now Equivant), is a risk assessment software used in the American criminal justice system to evaluate the likelihood of a defendant reoffending.
+Defendants taking the COMPAS test are given a questionnaire about topics ranging from family history to personal ideology.
+The questionnaire are then fed to the software system along with a number of parameters like the defendants's age, and then the system will assign a risk score to them from 1-10, with 10 being the highest risk.
 
-ProPublica is an American non-profit organization focused on investigative journalism in public interests. In 2016, they conduct a
+ProPublica is an American non-profit journalism organization focused on public interests.
+In 2016, they conduct an investigative report into the COMPAS system.
+They obtained the 2013-2014 COMPAS score data of over 10,000 defendants in Florida.
+They also obtained criminal records of these defendants through 2016 and compared if they actually recidivate or not.
+They only counted misdemeanors and felonies as recidivism but not less serious crimes such as infractions.
+In the study, they've found that black defendants are disproportionately scored higher than they actually are and white defendants are disproportionately scored lower than they actually are.
 
-\bye{
-fairness definition 1p
-    example
+We shall start applying our framework.
+There's generally no formal guide on how to set $\epsilon$, so we will take the 80\% rule's case and set it as $\epsilon = 0.2$.
+For the $\epsilon$ of positive and negative balance, we shall use $\epsilon = 5$ as it is a reasonable choice.
 
-framework definition 1p
-    abstraction
+Although this dataset scenario falls in the csv-as-input case in our framework, and thus the model $\model$ input isn't needed, we also trained two simple makeshift models using the dataset itself for demonstration that simply outputs binary $\mathtt{bool}$ prediction, ``will recidivate'' and ``will not recidivate'', and $\mathtt{int}$ risk score, $1$ to $10$.
+The model was trained by splitting the original dataset into 3 partitions: 60\% of the data was used for training; 20\% was used for validation; and the rest 20\% was used for calculating fairness measures.
+The accuracy of the binary model was about 60\% and the accuracy of the score model was about 25\%.
 
-framework example usage 1p
-    csv input
-    model input
+We will set the privileged predicate be to ``non-African-American'' so that the unprivileged group will be African-American to test if the African-American race is discriminated.
+\[
+    \priv(r_i) := r_i(\text{``race''}) \neq \text{``African-American''}
+\]
 
-compas introduction
+As ProPublica referenced Northpointe's COMPAS Practitioners Guide and cited that ``medium''(5-7) and ``high''(8-10) categories of scores are considered to indicate a risk of recidivism, we set the positive predicate using the readily available category.
+\[
+    \pos(r_i) := r_i(\text{``score\_text''}) \in \set{\text{``Medium''},\text{``High''}}
+\]
+Alternatively, since we know the corresponding scores of the categories, we can define it using the scores themselves, too.
+\[
+    \pos(r_i) := 5 \leq \mathtt{int}(r_i(\text{``decile\_score''})) \leq 10
+\]
+On the other hand, when we're taking prediction input from our makeshift model, we simply return its result because it already gives a binary $\mathtt{bool}$ result: $\pos(r_i) := \model(r_i)$
 
-compas analysis
-}
+For the ground truth predicate, the recidivism data is already present in the ProPublica dataset, so all we have to do is a simple lookup.
+\[
+    \tru(r_i) := r_i(\text{``two\_year\_recid''}) == \text{``1''}
+\]
+
+For the legitimate predicate, we may want to look at defendants with priors.
+\[
+    \leg(x)(r_i) := \mathtt{int}(r_i(\text{``priors\_count''})) > x
+\]
+
+For the score predicate, we may want to specifically look at risk scores between a specifically range.
+\[
+    \sco(u,l)(r_i) := l \leq \mathtt{int}(r_i(\text{``decile\_score''})) \leq u
+\]
+
+With these predicates defined, we can call the fairness measure functions and check if the measure hold or not.
+The results are compiled in Table~\ref{tab:results}.
+Since some of the fairness measures are equivalent, we write them in the same entry.
+
+\begin{table}[h]
+    \centering
+    \begin{tabular}{|l|c|c|} % 'l' specifies left alignment for the first column
+        \hline
+        \textbf{Fairness Measure} & \textbf{Criterion} & \textbf{Pass} \\
+        \hline
+Disparate Impact & 1.81 > 0.8 & YES \\
+\hline
+Demographic Parity & 0.26 < 0.2 & NO \\
+\hline
+Conditional Statistical Parity & 0.25 < 0.2 & NO \\
+\hline
+Overall Accuracy Equality & 0.02 < 0.2 & YES \\
+\hline
+Mean Difference & 0.26 < 0.2 & NO \\
+\hline
+\makecell[l]{Equalized Odds (true positive) \\ Equal Opportunity} & 0.23 < 0.2 & NO \\
+\hline
+\makecell[l]{Equalized Odds (false positive) \\ Predictive Equality} & 0.22 < 0.2 & NO \\
+\hline
+\makecell[l]{Conditional Use Accuracy Equality (true positive) \\ Predictive Parity} & 0.06 < 0.2 & YES \\
+\hline
+Conditional Use Accuracy Equality (true negative) & 0.06 < 0.2 & YES \\
+\hline
+Equal Calibration & 0.03 < 0.2 & YES \\
+\hline
+Positive Balance & 1.6 < 5 & YES \\
+\hline
+Negative Balance & 1.4 < 5 & YES \\
+\hline
+    \end{tabular}
+    \caption{Fairness measures of privileged group being non-African-American.}
+    \label{tab:results}
+\end{table}
+
+On first blush, it is curious that it satisfies disparate impact but not demographic parity.
+However, if we return to the definition, we'd see that disparate impact is meant to be used when being marked positive is an advantaged thing, while here in the COMPAS exmaple, being marked positive is a disadvantaged thing.
+Thus, henceforth we will exclude disparate impact from our analysis.
+
+We can then immediately tell from it failing demographic parity, mean difference, and conditional statistical parity that COMPAS prediction results was unfair against African-American, even if we only consider the ones with prior crimes.
+
+From the low overall accuracy equality and both conditional use accuracy equality criteria we can tell that the accuracy is similar across African-Americans and non-African-Americans.
+
+From failing both equalized odds criteria we can conclude that African-Americans are indeed treated unequally by the COMPAS system even after the ground truth is taken into account.
+This is the same conclusion reached by the ProPublica report.
+
+From the low equal calibration we can tell that if we only consider the medium risk score, African-Americans aren't treated unequally.
+
+Finally, if we look at positive and negative balance, we can see that they're of similar numbers. However, not much else can be said since the criterion $\epsilon = 5$ isn't actually justified.
 

### 2024-05-25 paper/abstract.tex
Added plt images

@@
 
-For the legitimate predicate, we may want to look at defendants with priors.
+For the legitimate predicate, we may want to look at defendants with priors. Here we shall set $args = (0)$.
 \[
@@
 
-For the score predicate, we may want to specifically look at risk scores between a specifically range.
+For the score predicate, we may want to specifically look at risk scores between a specifically range. Here we shall set $args = (7,5)$.
 \[
@@
 
-We can then immediately tell from it failing demographic parity, mean difference, and conditional statistical parity that COMPAS prediction results was unfair against African-American, even if we only consider the ones with prior crimes.
+We can then immediately tell from the failing demographic parity, mean difference, and conditional statistical parity that COMPAS prediction results was unfair against African-American, even if we only consider the ones with prior crimes.
 
@@
 
-From failing both equalized odds criteria we can conclude that African-Americans are indeed treated unequally by the COMPAS system even after the ground truth is taken into account.
+From the failing equalized odds we can conclude that African-Americans are indeed treated unequally by the COMPAS system even after the ground truth is taken into account.
 This is the same conclusion reached by the ProPublica report.
 
-From the low equal calibration we can tell that if we only consider the medium risk score, African-Americans aren't treated unequally.
+From the low equal calibration we can tell that if we only consider the medium risk score, African-Americans aren't treated fairly.
 

### 2024-05-26 paper/abstract.tex
Added pdf files

@@
 \usepackage{makecell}
+\usepackage{url}
+
+\graphicspath{{images/}}
+\DeclareGraphicsExtensions{.pdf}
 
@@
 
+%TODO package url
+
 \subsection{Application}
@@
     \end{tabular}
-    \caption{Fairness measures of privileged group being non-African-American.}
+    \caption{Fairness measures of privileged group being Non-African-American.}
     \label{tab:results}
@@
 
+Setting the privilege predicate to different races we can have a more comprehensive look over the dataset.
+We've checked the case of privilege predicate being Non-African-American, Non-Asian, Non-Caucasian, Non-Hispanic, and Non-Native-American.
+The results are shown in Figure~\ref{fig:compas}.
+
+From the results we can first notice that the overall accuracy eqaulity is low for both the Non-Asian and Non-Native-American case.
+This can be explained by checking the original dataset which shows that there are only 32 and 18 rows, respectivly, in dataset of 7214 rows.
+Hence, the accuracy is naturally lower because of the small data size.
+
+\begin{figure*}[h]
+    \centering
+    \begin{minipage}[b]{0.48\linewidth}
+        \centering
+        \includegraphics[width=\linewidth]{Non-African-American}
+        \subcaption{Privileged Group: Non-African-American}\label{fig:1}
+    \end{minipage}
+    \hfill
+    \begin{minipage}[b]{0.48\linewidth}
+        \centering
+        \includegraphics[width=\linewidth]{Non-Asian}
+        \subcaption{Privileged Group: Non-Asian}\label{fig:2}
+    \end{minipage}
+    \hfill
+    \begin{minipage}[b]{0.48\linewidth}
+        \centering
+        \includegraphics[width=\linewidth]{Non-Caucasian}
+        \subcaption{Privileged Group: Non-Caucasian}\label{fig:2}
+    \end{minipage}
+    \hfill
+    \begin{minipage}[b]{0.48\linewidth}
+        \centering
+        \includegraphics[width=\linewidth]{Non-Hispanic}
+        \subcaption{Privileged Group: Non-Hispanic}\label{fig:2}
+    \end{minipage}
+    \hfill
+    \begin{minipage}[b]{0.48\linewidth}
+        \centering
+        \includegraphics[width=\linewidth]{Non-Native-American}
+        \subcaption{Privileged Group: Non-Native-American}\label{fig:2}
+    \end{minipage}
+    \caption{Fairness Measures of ProPublica COMPAS dataset}
+    \label{fig:compas}
+\end{figure*}
+
 \bibliographystyle{IEEEtran}

### 2024-05-26 paper/abstract.tex
Added balance plot

@@
 
-%TODO package url
+We implemented the framework in Python and it is available as an public domain open-sourced Python package at \url{https://pypi.org/project/fairness-checker/}.
 
@@
 Setting the privilege predicate to different races we can have a more comprehensive look over the dataset.
-We've checked the case of privilege predicate being Non-African-American, Non-Asian, Non-Caucasian, Non-Hispanic, and Non-Native-American.
+We've checked the case of privilege predicate being Non-African-American, Non-Asian, Non-Caucasian, Non-Hispanic, and Non-Native American.
 The results are shown in Figure~\ref{fig:compas}.
 
-From the results we can first notice that the overall accuracy eqaulity is low for both the Non-Asian and Non-Native-American case.
+From the results we can first notice that the overall accuracy eqaulity is low for both the Non-Asian and Non-Native American case.
 This can be explained by checking the original dataset which shows that there are only 32 and 18 rows, respectivly, in dataset of 7214 rows.
 Hence, the accuracy is naturally lower because of the small data size.
+
+Looking at the remaining three groups, Non-African-American, Non-Caucasian, and Non-Hispanic, their accuracy are relatively much better.
+
+As for fairness measures, we can immediately see from that all measures are $\epsilon < 0.2$ for Non-Caucasian and Non-Hispanic that African-American are indeed treated unfairly in a broad sense.
+
+More specifically, African-American are treated more unfairly according to demographic parity, conditional statistical parity, mean difference, and equalized odds.
+This is again in accordance with the conclusion of the ProPublica report.
 
@@
         \centering
+        \includegraphics[width=\linewidth]{Non-Caucasian}
+        \subcaption{Privileged Group: Non-Caucasian}\label{fig:2}
+    \end{minipage}
+    \hfill
+    \begin{minipage}[b]{0.48\linewidth}
+        \centering
         \includegraphics[width=\linewidth]{Non-Asian}
@@
         \centering
-        \includegraphics[width=\linewidth]{Non-Caucasian}
-        \subcaption{Privileged Group: Non-Caucasian}\label{fig:2}
+        \includegraphics[width=\linewidth]{Non-Native American}
+        \subcaption{Privileged Group: Non-Native American}\label{fig:2}
     \end{minipage}
@@
         \centering
-        \includegraphics[width=\linewidth]{Non-Native-American}
-        \subcaption{Privileged Group: Non-Native-American}\label{fig:2}
+        \includegraphics[width=\linewidth]{race balance}
+        \subcaption{Balance}\label{fig:2}
     \end{minipage}
-    \caption{Fairness Measures of ProPublica COMPAS dataset}
+    \caption{Fairness Measures of ProPublica COMPAS dataset grouped by race}
     \label{fig:compas}

### 2024-05-27 paper/abstract.tex
Added all plots
Added placeins

@@
 \usepackage{url}
+\usepackage{placeins}
 
@@
 
+On the other hand, we also analyzed the case of unprivileged group being across the three age groups and the case of privileged group of sex being Male and charged degree being misdemeanor in Figure~\ref{fig:compas2}.
+
+We can see that, interestingly, the group of age ``25 - 45'' group receives generally fair treatment.
+Its data size is also the largest at 4109 rows, whereas ``Less than 25'' age group has 1529 rows and ``Greater than 45'' age group has 1576 rows, so the possibility of skewed data is unlikely.
+
+If we look closer we can see that in the age group ``Less than 25'', its predictive equality is noticeably worse, meaning in young people false positive rate is higher than older people; conversely, when we look at the age group ``Greater than 45'', both equal opportunity and predictive equality are worse, meaning for old people their treatment is even more unfair than young people.
+Only the middle age group is treated fairly.
+
+In another vein, the case of ``Male'' group and ``Misdemeanor'' group are both treated rather fairly.
+Furthermore, their overeal accuracy equality are both extremely good at the $< 0.1$ range.
+The predictive equality  of ``Male'' group even has a measure as low as $0.000004$.
+
 \begin{figure*}[h]
@@
     \end{minipage}
-    \caption{Fairness Measures of ProPublica COMPAS dataset grouped by race}
+    \caption{Fairness measures of ProPublica COMPAS dataset grouped by race}
     \label{fig:compas}
 \end{figure*}
+
+\begin{figure*}[h]
+    \centering
+    \begin{minipage}[b]{0.48\linewidth}
+        \centering
+        \includegraphics[width=\linewidth]{Non-Less than 25}
+        \subcaption{Privileged Group: Non-Less than 25}\label{fig:1}
+    \end{minipage}
+    \hfill
+    \begin{minipage}[b]{0.48\linewidth}
+        \centering
+        \includegraphics[width=\linewidth]{Male}
+        \subcaption{Privileged Group: Male}\label{fig:2}
+    \end{minipage}
+    \hfill
+    \begin{minipage}[b]{0.48\linewidth}
+        \centering
+        \includegraphics[width=\linewidth]{Non-25 - 45}
+        \subcaption{Privileged Group: Non-25 - 45}\label{fig:2}
+    \end{minipage}
+    \hfill
+    \begin{minipage}[b]{0.48\linewidth}
+        \centering
+        \includegraphics[width=\linewidth]{Misdemeanor}
+        \subcaption{Privileged Group: Misdemeanor}\label{fig:2}
+    \end{minipage}
+    \hfill
+    \begin{minipage}[b]{0.48\linewidth}
+        \centering
+        \includegraphics[width=\linewidth]{Non-Greater than 45}
+        \subcaption{Privileged Group: Non-Greater than 45}\label{fig:2}
+    \end{minipage}
+    \hfill
+    \begin{minipage}[b]{0.48\linewidth}
+        \centering
+        \includegraphics[width=\linewidth]{misc balance}
+        \subcaption{Balance}\label{fig:2}
+    \end{minipage}
+    \caption{Fairness measures of ProPublica COMPAS dataset grouped by age, sex, and charged degree}
+    \label{fig:compas2}
+\end{figure*}
+
+\FloatBarrier
 

### 2024-05-27 paper/paper.tex
Renamed

--- 2024-05-27 paper/abstract.tex
@@
 
+\section{Conclusion}
+Addressing the challenges brought upon by the rapid advancements in AI technologies and its introduced bias necessitates a rigorous assessment of AI system fairness.
+While fairness is subjective and, it can be defined through various statistical measures.
+Our research contributes to this effort by proposing a comprehensive framework for auditing AI systems using multiple fairness metrics, such as demographic parity, equalized odds, and overall accuracy equality.
+
+By applying this framework to the COMPAS dataset, we confirmed that African-American defendants were unfairly scored compared to other racial groups.
+These results align with ProPublica's findings, demonstrating the utility and accuracy of our fairness auditing tool.
+Developed in Python and publicly available, this tool enables third parties to conduct detailed assessments of AI systems, promoting transparency and accountability.
+
 \begin{figure*}[h]

### 2024-05-28 paper/paper.tex
Added todos

@@
 \newcommand{\model}{\mathcal{M}}
-\newcommand{\priv}{P}
+\newcommand{\priv}{R}
 \newcommand{\pos}{\hat{P}}
@@
 
+%TODO emph
+Protected attributes are the characteristics of individuals that are, for example, legally or ethically, considered sensitive and warrant protection against discrimination and bias.
 We denote by $S$ some protected attribute.
-Protected attributes are the characteristics of individuals that are, for example, legally or ethically, considered sensitive and warrant protection against discrimination and bias.
 We write $S = 1$ to represent the privileged group and $S \neq 1$ to represent the unprivileged group.
@@
 
-\subsection{Preliminaries}
+\subsection{Notations}
+%TODO add function type alpha to beta
+%TODO replace str -> str to key -> value concretely theyre strings
+%TODO replace key -> value with row type
+%TODO new definite use emph
 Let $\db = \set{r_1, r_2, ...}$ be a database containing rows of data.
@@
 A positive predicate of rows $\pos : (\mathtt{str} \implies \mathtt{str}) \implies \mathtt{bool}$ takes a row and determines if its prediction is positive.
-For example, $\pos(r_i) := \mathtt{int}(r_i(\text{``score''})) > \text{``7''}$ means a row's prediction is positive if its score is greater than 7.
+For example, $\pos(r_i) := \mathtt{int}(r_i(\text{``score''})) > 7$ means a row's prediction is positive if its score is greater than 7.
 A positive predicate of model results $\pos : \alpha \implies \mathtt{bool}$ takes a model's result and determines if it is positive.
@@
 Given a datasets for auditing use, and given csv prediction results or the prediction model itself, we can calculate the fairness measures mentioned in Section~\ref{sec:measures} by:
+
+%TODO: explain priv pos relationship with Y S
 
@@
 
-\subsection{Application}
+\section{Application}
+\subsection{Setup}
 \label{sec:app}
@@
 
+\subsection{Fairness Analysis of African-American group}
 On first blush, it is curious that it satisfies disparate impact but not demographic parity.
@@
 
+%TODO: on a scale of 10, we have 1.5  difference on avg, and it remains to be said that if this is fair, should consult expert by letting them choose epsilon
+%TODO: remove epsilon=5
 Finally, if we look at positive and negative balance, we can see that they're of similar numbers. However, not much else can be said since the criterion $\epsilon = 5$ isn't actually justified.
 
+\subsection{Fairness Analysis of Different Races}
 Setting the privilege predicate to different races we can have a more comprehensive look over the dataset.
@@
 
+\subsection{Fairness Analysis of Different Groups}
 On the other hand, we also analyzed the case of unprivileged group being across the three age groups and the case of privileged group of sex being Male and charged degree being misdemeanor in Figure~\ref{fig:compas2}.
@@
 \section{Conclusion}
+%TODO: mention we're white box
 Addressing the challenges brought upon by the rapid advancements in AI technologies and its introduced bias necessitates a rigorous assessment of AI system fairness.

### 2024-05-29 setup.py
Changed wording on calibration

@@ -5,7 +5,7 @@ with open("src/README.md", "r") as f:
 
 setup(
     name="fairness_checker",
-    version="0.0.1",
+    version="0.0.2",
     package_dir={'': 'src'},
     packages=find_packages(where="src"),
     description="Fairnes checker",

### 2024-05-29 src/fairness_checker/main.py

@@ -295,8 +295,8 @@ class fairness_model_checker:
                           ratio: float,
                           model: object,
                           privileged_predicate: Callable[[csv_row], bool],
-                          score_predicate_h: Callable[..., Callable[[csv_row], bool]],
-                          score_arg: Tuple[Any, ...],
+                          calib_predicate_h: Callable[..., Callable[[csv_row], bool]],
+                          calib_arg: Tuple[Any, ...],
                           truth_predicate: Callable[[csv_row], bool]) -> bool:
         """
         Evaluates the equal calibration of the model's predictions between privileged and unprivileged groups.
@@ -320,10 +320,10 @@ class fairness_model_checker:
                 privileged_Y_pred = model.predict('privileged.csv')
                 unprivileged_Y_pred = model.predict('unprivileged.csv')
 
-        score_predicate = score_predicate_h(*score_arg)
+        calib_predicate = calib_predicate_h(*calib_arg)
 
-        privileged_score = list(filter(lambda row_Y: score_predicate(row_Y[1]), zip(privileged, privileged_Y_pred)))
-        unprivileged_score = list(filter(lambda row_Y: score_predicate(row_Y[1]), zip(unprivileged, unprivileged_Y_pred)))
+        privileged_score = list(filter(lambda row_Y: calib_predicate(row_Y[1]), zip(privileged, privileged_Y_pred)))
+        unprivileged_score = list(filter(lambda row_Y: calib_predicate(row_Y[1]), zip(unprivileged, unprivileged_Y_pred)))
 
         privileged_score = list(map(lambda row_Y: row_Y[0], privileged_score))
         unprivileged_score = list(map(lambda row_Y: row_Y[0], unprivileged_score))
@@ -819,8 +819,8 @@ class fairness_csv_checker:
     def equal_calibration(self,
                           ratio: float,
                           privileged_predicate: Callable[[csv_row], bool],
-                          score_predicate_h: Callable[..., Callable[[csv_row], bool]],
-                          score_arg: Tuple[Any, ...],
+                          calib_predicate_h: Callable[..., Callable[[csv_row], bool]],
+                          calib_arg: Tuple[Any, ...],
                           truth_predicate: Callable[[csv_row], bool]) -> bool:
         """
         Evaluates the equal calibration of the model's predictions between privileged and unprivileged groups.
@@ -834,10 +834,10 @@ class fairness_csv_checker:
         Returns:
[…]

### 2024-05-29 paper/paper.tex
Revised version

@@
 \newcommand{\sco}{\hat{S}}
+\newcommand{\calib}{C}
 
@@
 
-%TODO emph
-Protected attributes are the characteristics of individuals that are, for example, legally or ethically, considered sensitive and warrant protection against discrimination and bias.
+\emph{Protected attributes} are the characteristics of individuals that are, for example, legally or ethically, considered sensitive and warrant protection against discrimination and bias.
 We denote by $S$ some protected attribute.
-We write $S = 1$ to represent the privileged group and $S \neq 1$ to represent the unprivileged group.
+We write $S = 1$ to represent the \emph{privileged group} and $S \neq 1$ to represent the \emph{unprivileged group}.
 For example, let $S$ be Caucasian.
@@
 \subsection{Notations}
-%TODO add function type alpha to beta
-%TODO replace str -> str to key -> value concretely theyre strings
-%TODO replace key -> value with row type
-%TODO new definite use emph
-Let $\db = \set{r_1, r_2, ...}$ be a database containing rows of data.
-Each row $r_i : \mathtt{str} \implies \mathtt{str}$ is a lookup table or dictionary.
+Types $\alpha,\beta$ are stand-ins for any type.
+Let $\alpha \implies \beta$ denote a function from type $\alpha$ to type $\beta$.
+$\alpha^n$ denotes multiple instances of type $\alpha$; $\alpha_i$ denotes some particular type; $\alpha_i^n$ denotes multiple instances of possibly different types.
+We write $x : \alpha$ to mean $x$ is of the type $\alpha$.
+
+A \emph{database} $\db = \set{r_1, r_2, ...}$ is a collection of rows.
+A \emph{row} $r_i : \mathtt{key} \implies \mathtt{value}$ is a lookup table or dictionary, where the concrete type of $\mathtt{key}$ and $\mathtt{value}$ is $\mathtt{str}$.
 For example, $r_n(\text{``sex''}) = \text{``Female''}$ means $r_n$'s sex is female.
+Henceforth we will use $\mathtt{row}$ as a type synonym of $\mathtt{key} \implies \mathtt{value}$.
 
-A model $\model : (\mathtt{str} \implies \mathtt{str}) \implies \alpha$ takes a row and returns the model's prediction result of some type $\alpha$.
+A \emph{model} $\model : \mathtt{row} \implies \alpha$ is a black-box predictor that takes a row and returns the its prediction result of some type $\alpha$; for example, if the model returns a string then $\alpha = \mathtt{str}$.
 
-A privileged predicate $\priv : (\mathtt{str} \implies \mathtt{str}) \implies \mathtt{bool}$ takes a row and determines if it belongs to the privileged group.
+A \emph{privileged predicate} $\priv : \mathtt{row} \implies \mathtt{bool}$ takes a row and determines if it belongs to the privileged group.
 For example, $\priv(r_i) := r_i(\text{``race''}) == \text{``Caucasian''}$ means the privileged group is those with race being Caucasian.
 
-A positive predicate of rows $\pos : (\mathtt{str} \implies \mathtt{str}) \implies \mathtt{bool}$ takes a row and determines if its prediction is positive.
+A \emph{positive predicate} of rows $\pos : \mathtt{row} \implies \mathtt{bool}$ takes a row and determines if its prediction is positive.
 For example, $\pos(r_i) := \mathtt{int}(r_i(\text{``score''})) > 7$ means a row's prediction is positive if its score is greater than 7.
@@
 
-A ground truth predicate $\tru : (\mathtt{str} \implies \mathtt{str}) \implies \mathtt{bool}$ takes a row and gives the ground truth of the result.
+A \emph{score predicate} of rows $\sco : \mathtt{row} \implies \alpha$ takes a row and gives its predicted score.
+Similar to positive predicate, a score predicate of model results $\sco : \alpha \implies \beta$ takes a model's result and gives its score.
 
-A legitimate predicate $\leg : \alpha_i^n \implies (\mathtt{str} \implies \mathtt{str}) \implies \mathtt{bool}$, and similarly a score predicate $\sco : \alpha_i^n \implies (\mathtt{str} \implies \mathtt{str}) \implies \mathtt{bool}$, takes $n$ parameters and returns a row predicate.
-For example, $\sco(u, l)(r_i) := l < \mathtt{int}(r_i(\text{``score''})) < u$ first takes two parameters $u,l$ as upper bound and lower bound, and then decides if $r_i$'s prediction score is between them.
+A \emph{ground truth predicate} $\tru : \mathtt{row} \implies \mathtt{bool}$ takes a row and gives the ground truth of the result.
+
+A \emph{legitimate predicate} $\leg : \alpha_i^n \implies \mathtt{row} \implies \mathtt{bool}$ takes $n$ parameters and returns a row predicate.
+For example, $\leg(x)(r_i) := \mathtt{int}(r_i(\text{``priors\_count''})) > x$ first take a parameter $x$ and then decides if priors count is larger than it.
+
+A \emph{calibration predicate} $\calib : \alpha_i^n \implies \mathtt{row} \implies \mathtt{bool}$, takes $n$ parameters and returns a row predicate.
+For example, $\calib(u, l)(r_i) := l < \mathtt{int}(r_i(\text{``score''})) < u$ first takes two parameters $u,l$ as upper bound and lower bound, and then decides if $r_i$'s prediction score is between them.
 
@@
 
-Given a datasets for auditing use, and given csv prediction results or the prediction model itself, we can calculate the fairness measures mentioned in Section~\ref{sec:measures} by:
+Given a datasets $\db$ for auditing use, and given csv prediction results or the prediction model $\model$ itself, we can calculate the fairness measures by modeling the statistical random variables mentioned in Section~\ref{sec:measures} with our predicates.
 
-%TODO: explain priv pos relationship with Y S
+For a simple exmaple, for equal opportunity, recall its formal definition.
+\begin{align*}
+    \abs{P[\hat{Y} = 1 | S = 1, Y = 1] - P[\hat{Y} = 1 | S \neq 1, Y = 1]} \leq \epsilon
+\end{align*}
+To model the $Y$, $\hat{Y}$, and $S$, we we define the corresponding $\tru$, $\pos$, and $\priv$.
+We have $Y = 1$ if and only if $\tru$ is true; $\hat{Y} = 1$ if and only if $\pos$ is true; $S = 1$ if and only if $\priv$ is true; and vice versa.
+\begin{align*}
+    \text{equal\_opportunity} (\epsilon, \model, \priv, \pos, \tru)
+\end{align*}
+
+For positive balance, recall its formal definition.
+\begin{align*}
+    \abs{E[\hat{V} | Y = 0, S = 1] - E[\hat{V} | Y = 0, S \neq 1]} \leq \epsilon
+\end{align*}
+We model $Y$ and $S$ as the previous example. As for $\hat{V}$ we have it equal to the output of $\sco$, and now we can calculate the measure.
+\begin{align*}
+    \text{positive\_balance} (\epsilon, \model, \priv, \sco, \tru)
+\end{align*}
+
+For another exmaple, recall the formal definition of equal calibration.
+\begin{align*}
+    \abs{P[Y = 1 | S = 1, \hat{V} = v] - P[Y = 1 | S \neq 1, \hat{V} = v]} \leq \epsilon
+\end{align*}
+Here we model $Y$ and $S$ as above. As for $\hat{V} = v$, we abstracted $\hat{V}$ with $\calib$ and $v$ with $args$. We have $\hat{V} = v$ if and only if $\calib(args, ...)$ is true.
+\begin{align*}
+    \text{equal\_calibration} (\epsilon, \model, \priv, \calib, \tru, (args, ...))
+\end{align*}
+
+For conditional statistical parity, its formal definition is:
+\[
+    \abs{P[\hat{Y} = 1 | S = 1, L = l] - P[\hat{Y} = 1 | S \neq 1, L = l]} \leq \epsilon
+\]
+$\hat{Y}$ and $S$ are modeled as above. As for $L = l$, we abstracted it similarly to the case above by having $L = l$ if and only if $\leg(args, ...)$ is true.
+\begin{align*}
+    \text{conditional\_statistical\_parity} (\epsilon, \model, \priv, \pos, \leg, (args, ...))
+\end{align*}
+
+This way, we can calculate each conditions programmatically by testing if the predicate is true on all rows and obtain the resulting measure.
+All the measures can be modeled in a similar vein.
+So, in total, we have:
 
@@
 \text{predictive\_parity} & (\epsilon, \model, \priv, \pos, \tru) \\
-\text{equal\_calibration} & (\epsilon, \model, \priv, \sco, \tru, (args, ...)) \\
-\text{positive\_balance} & (\epsilon, \model, \priv, \sco, \tru, (args, ...)) \\
-\text{nagative\_balance} & (\epsilon, \model, \priv, \sco, \tru, (args, ...)) \\
+\text{equal\_calibration} & (\epsilon, \model, \priv, \calib, \tru, (args, ...)) \\
+\text{positive\_balance} & (\epsilon, \model, \priv, \sco, \tru) \\
+\text{nagative\_balance} & (\epsilon, \model, \priv, \sco, \tru) \\
 \end{align*}
@@
 \[
-    \leg(x)(r_i) := \mathtt{int}(r_i(\text{``priors\_count''})) > x
+    \leg(0)(r_i) := \mathtt{int}(r_i(\text{``priors\_count''})) > 0
 \]
 
-For the score predicate, we may want to specifically look at risk scores between a specifically range. Here we shall set $args = (7,5)$.
+For the calibration predicate, we may want to specifically look at risk scores between a specifically range. Here we shall set $args = (7,5)$.
 \[
-    \sco(u,l)(r_i) := l \leq \mathtt{int}(r_i(\text{``decile\_score''})) \leq u
+    \calib(7,5)(r_i) := 5 \leq \mathtt{int}(r_i(\text{``decile\_score''})) \leq 7
 \]
@@
 \hline
-Positive Balance & 1.6 < 5 & YES \\
+Positive Balance & 1.6 & ? \\
 \hline
-Negative Balance & 1.4 < 5 & YES \\
+Negative Balance & 1.4 & ? \\
 \hline
@@
 
-%TODO: on a scale of 10, we have 1.5  difference on avg, and it remains to be said that if this is fair, should consult expert by letting them choose epsilon
-%TODO: remove epsilon=5
-Finally, if we look at positive and negative balance, we can see that they're of similar numbers. However, not much else can be said since the criterion $\epsilon = 5$ isn't actually justified.
+Finally, if we look at positive and negative balance, we can see that they're of similar numbers.
+Their average is 1.5 on a scale of 10, which means approximately a 15\% difference.
+It remains to be said if this is fair or not.
+An auditor could consult a domain expert for advice on how to set a $\epsilon$ and how is it justified.
 
@@
 \section{Conclusion}
-%TODO: mention we're white box
 Addressing the challenges brought upon by the rapid advancements in AI technologies and its introduced bias necessitates a rigorous assessment of AI system fairness.
 While fairness is subjective and, it can be defined through various statistical measures.
-Our research contributes to this effort by proposing a comprehensive framework for auditing AI systems using multiple fairness metrics, such as demographic parity, equalized odds, and overall accuracy equality.
+Our research contributes to this effort by proposing a comprehensive framework for auditing AI systems using multiple white-box fairness metrics, such as demographic parity, equalized odds, and overall accuracy equality.
 
@@
 
-\FloatBarrier
+%\FloatBarrier
 

### 2024-06-04 paper/paper.tex
Some minor formatting change
Added todos

@@
 
-We denote by $\epsilon$ some threshold that's used to limit the fairness measures.
+We denote by $\epsilon$ some threshold that is used to limit the fairness measures.
+
+%TODO: explain various measures have different origins.
+for presentation we assume Y=1 to be detrimental.
+use comaps as example and the intuition is Y=1 => reoffend; but if the case is school or loan the tuition is reversed
 
@@
 Formally \cite{feldman2015certifying}, it is:
-
 \[
@@
 \]
-
 In the case of 80\% rule, $\epsilon = 20\%$.
@@
 Formally, it is:
-
 \[
@@
 \]
-
 where $L$ denotes the legitimate attributes.
@@
 \subsection{Mean Difference}
+%TODO: rewrite, move compas example to positive balance
 Mean difference \cite{vzliobaite2017measuring} considers the case of non-binary outcome, such as the score of COMPAS. It takes the expected value of the prediction. Formally, it is:
-
 \[
@@
 Formally, it is:
-
 \begin{align*}
@@
 \subsection{Equal Opportunity}
+%TODO: rewrite use footnote to explain name origin
 Equal opportunity \cite{hardt2016equality} is a relaxation of equalized odds by only considering the true positive case.
@@
 \subsection{Predictive Parity}
+%TODO: rewrite, set positive = detrimental to all cases
 Predictive parity \cite{chouldechova2017fair} is a relaxation of conditional use accuracy equality by only considering the positive predictive value case.
@@
 Negative balance \cite{kleinberg2016inherent} is like positive balance except it conditions on the case of $Y = 0$. Formally, it is:
-
 \begin{align*}
@@
 We shall start applying our framework.
-There's generally no formal guide on how to set $\epsilon$, so we will take the 80\% rule's case and set it as $\epsilon = 0.2$.
+There is generally no formal guide on how to set $\epsilon$, so we will take the 80\% rule's case and set it as $\epsilon = 0.2$.
 For the $\epsilon$ of positive and negative balance, we shall use $\epsilon = 5$ as it is a reasonable choice.
 
+%TODO: move this to the back and choose one measure to comapre with compas
 Although this dataset scenario falls in the csv-as-input case in our framework, and thus the model $\model$ input isn't needed, we also trained two simple makeshift models using the dataset itself for demonstration that simply outputs binary $\mathtt{bool}$ prediction, ``will recidivate'' and ``will not recidivate'', and $\mathtt{int}$ risk score, $1$ to $10$.
@@
 \]
+
+%TODO: for score predicate
 

### 2024-06-04 setup.py
Added python requires

@@ -5,7 +5,7 @@ with open("src/README.md", "r") as f:
 
 setup(
     name="fairness_checker",
-    version="0.0.2",
+    version="0.0.3",
     package_dir={'': 'src'},
     packages=find_packages(where="src"),
     description="Fairnes checker",
@@ -15,4 +15,5 @@ setup(
     author="RexYuan",
     author_email="hello@rexyuan.com",
     license="Unlicense",
+    python_requires=">=3.6",
 )

### 2024-06-04 paper/paper.tex
Major revision

--- 2024-05-29 paper/paper.tex
@@
 Cases such as COMPAS underscore the critical importance of rigorously examining the fairness of AI systems by third parties.
-Fairness is fundementally a subjective social construct, heavily influenced by cultural context and deeply rooted in historical inequalities.
+Fairness is fundamentally a subjective social construct, heavily influenced by cultural context and deeply rooted in historical inequalities.
 However, there have been development in research leveraging statistical metrics to quantify fairness that provide transparent and objective insights.
 
-By employing such mathematically rigorous methodologies, our research can offer a more deeper look into the fairness of AI systems, enabling third parties to make more informed and robust judgment.
+By employing such mathematically rigorous methodologies, our research can offer a deeper look into the fairness of AI systems, enabling third parties to make more informed and robust judgment.
 For example, it would reveal that the COMPAS system violates a fairness measure called equalized odds.
@@
 
-Our tool is written in Python and is offered as a Python package. It supports common datasets format such as csv. It is open-sourced and publicly available for downloads.
+Our tool is written in Python and is offered as a Python package. It supports common datasets format such as CSV. It is open-sourced and publicly available for downloads.
 
@@
 To have a constructive discussion on fairness, we must first have precise definitions of it.
-Pessach and Shmueli \cite{pessach2022review} formulated a number of fairness measures in an unified mathematical notation.
+Pessach and Shmueli \cite{pessach2022review} formulated a number of fairness measures in a unified mathematical notation.
 We will base our framework on their formulation.
@@
 
-We denote by $\epsilon$ some threshold that's used to limit the fairness measures.
+We denote by $\epsilon$ some threshold that is used to limit the fairness measures.
+
+A positive outcome or prediction may be either beneficial or detrimental to the individual in different cases.
+Various measures were devised when considering different scenarios.
+For presentation, we consider $Y=1$ to be the detrimental case, like in the COMPAS prediction scenarios.
 
 \subsection{Disparate Impact}
-In 1971 \cite{griggs1971}, the US spreme court ruled that it is illegal for hiring decisions to have  ``disparate impact'' by race, thus coining the term.
+In 1971 \cite{griggs1971}, the US Supreme Court ruled that it is illegal for hiring decisions to have  ``disparate impact'' by race, thus coining the term.
 It is taken as unintentional discrimination, as opposed to intentional discrimination, which is called ``disparate treatment''.
@@
 \subsection{Overall Accuracy Equality}
-Overall accuracy eqaulity \cite{berk2021fairness} is similar to demographic parity, but instead of the case of $\hat{Y} = 1$, it considers the case of $Y = \hat{Y}$; that is, the case where the prediction is accurate.
+Overall accuracy equality \cite{berk2021fairness} is similar to demographic parity, but instead of the case of $\hat{Y} = 1$, it considers the case of $Y = \hat{Y}$; that is, the case where the prediction is accurate.
 Formally, it is:
-
 \[
@@
 \subsection{Mean Difference}
-Mean difference \cite{vzliobaite2017measuring} considers the case of non-binary outcome, such as the score of COMPAS. It takes the expected value of the prediction. Formally, it is:
-
+Mean difference \cite{vzliobaite2017measuring} considers the expected value of the prediction. Formally, it is:
 \[
@@
 \subsection{Equalized Odds}
-Equalized odds \cite{hardt2016equality} is similar to demogrphic parity, but it further takes into account the grouth truth.
-It consider the cases of true positive and false positive.
+Equalized odds \cite{hardt2016equality} is similar to demographic parity, but it further takes into account the ground truth.
+It considers the cases of true positive and false positive.
 It solves the downsides of a fully accurate classifier might be deemed unfair by measures that don't take ground truth into consideration.
-For example, consider a group $A$ that is predicated to recidivate and they do in fact always recidivate and a group $B$ that is predicated to recidivate but end up never recidivating.
-A fully accurate classifier will always predict $A$ to recidivate while $B$ to never recidivate, and this will violate demogrphic parity.
+For example, consider a group $A$ that is predicated to recidivate and they do in fact recidivate and a group $B$ that is predicated to recidivate but end up never recidivating.
+A fully accurate classifier will always predict $A$ to recidivate while $B$ to never recidivate, and this will violate demographic parity.
 By taking ground truth into account, equalized odds avoids these pitfalls.
 Formally, it is:
-
 \begin{align*}
@@
 \subsection{Equal Opportunity}
-Equal opportunity \cite{hardt2016equality} is a relaxation of equalized odds by only considering the true positive case.
-It is named so because when outcome being positive is beneficial to the individual, as it is an opportunity, such as school acceptance, the fairness of the true positive case is much more important.
+Equal opportunity \footnote{It was named so by considering the cases where positive outcomes benefit individuals, such as school admissions, hence the word opportunity.} \cite{hardt2016equality} is a relaxation of equalized odds by only considering the true positive case.
 Formally, it is:
-
 \begin{align*}
@@
 \subsection{Predictive Equality}
-Predictive equality \cite{corbett2017algorithmic} is also a relaxation of equalized odds by only considering the false positive case.
-It was formulated by considering positive outcome being detrimental to the individual, such as predicting future recidivism.
-Thus, only the false positive case is taken.
-Formally, it is:
-
+Predictive equality \cite{corbett2017algorithmic} is also a relaxation of equalized odds by only considering the false positive case. Formally, it is:
 \begin{align*}
@@
 Conditional use accuracy equality \cite{berk2021fairness} is similar to equalized odds, but instead of conditioning on the ground truth, it conditions on the prediction and calculated the probability of the ground truth.
-It can be seem as checking the prediction accuracy across groups, thus the name.
+It can be seen as checking the prediction accuracy across groups, thus the name.
 It further requires the measure on the case of positive predictive values is less than that of the negative predictive values.
 Formally, it is:
-
 \begin{align*}
@@
 \subsection{Predictive Parity}
-Predictive parity \cite{chouldechova2017fair} is a relaxation of conditional use accuracy equality by only considering the positive predictive value case.
-Like predictive equality, it was formulated by considering positive outcome being detrimental to the individual.
-Thus, only the positive predictive value case is taken.
-Formally, it is:
-
+Predictive parity \cite{chouldechova2017fair} is a relaxation of conditional use accuracy equality by only considering the positive predictive value case. Formally, it is:
 \begin{align*}
@@
 \subsection{Positive Balance}
-Positive balance \cite{kleinberg2016inherent} is similar to equal opportunity, but instead of taking the difference of probability of binary prediction $\hat{Y}$, it takes the difference of the expected value of the score $\hat{V}$.
+Positive balance \cite{kleinberg2016inherent} is similar to equal opportunity, but instead of taking the difference of probability of binary prediction $\hat{Y}$, it takes the difference of the expected value of the score $\hat{V}$, which may be non-binary, such as the score of COMPAS.
 Formally, it is:
-
 \begin{align*}
@@
 \hline
-Overall Accuracy Eqaulity & $\abs{P[Y = \hat{Y} | S = 1] - P[Y = \hat{Y} | S \neq 1]} \leq \epsilon$ \\
+Overall Accuracy Equality & $\abs{P[Y = \hat{Y} | S = 1] - P[Y = \hat{Y} | S \neq 1]} \leq \epsilon$ \\
 \hline
@@
 Per the review by Pessach and Shmueli \cite{pessach2022review}, we designed an auditing framework for calculating the various fairness measure.
-We offer two versions of fairness checkers: one for when the prediction positive results are readily available in csv input and one for when a model is provided.
-In most auditing cases the model version is preferred because csv result can be easily fabricated.
+We offer two versions of fairness checkers: one for when the prediction positive results are readily available in CSV input and one for when a model is provided.
+In most auditing cases the model version is preferred because CSV result can be easily fabricated.
 
@@
 Let $\alpha \implies \beta$ denote a function from type $\alpha$ to type $\beta$.
-$\alpha^n$ denotes multiple instances of type $\alpha$; $\alpha_i$ denotes some particular type; $\alpha_i^n$ denotes multiple instances of possibly different types.
+$\alpha_i$ denotes some particular type; $\prod_i \alpha_i$ denotes the Cartesian product of multiple $\alpha$ of possibly different types.
 We write $x : \alpha$ to mean $x$ is of the type $\alpha$.
+We write $\mathtt{str}$ for the string type, $\mathtt{bool}$ for the boolean type, and $\mathtt{int}$ for the integer type.
 
@@
 For example, $r_n(\text{``sex''}) = \text{``Female''}$ means $r_n$'s sex is female.
-Henceforth we will use $\mathtt{row}$ as a type synonym of $\mathtt{key} \implies \mathtt{value}$.
+Henceforth, we will use $\mathtt{row}$ as a type synonym of $\mathtt{key} \implies \mathtt{value}$.
 
-A \emph{model} $\model : \mathtt{row} \implies \alpha$ is a black-box predictor that takes a row and returns the its prediction result of some type $\alpha$; for example, if the model returns a string then $\alpha = \mathtt{str}$.
+A \emph{model} $\model : \mathtt{row} \implies \alpha$ is a black-box predictor that takes a row and returns its prediction result of some type $\alpha$; for example, if the model returns a string then $\alpha = \mathtt{str}$.
 
@@
 A positive predicate of model results $\pos : \alpha \implies \mathtt{bool}$ takes a model's result and determines if it is positive.
-In the simplest case, the model returns a $\mathtt{bool}$, and the positive predicate can just be the idnetity function.
+In the simplest case, the model returns a $\mathtt{bool}$, and the positive predicate can just be the identity function.
 
@@
 
-A \emph{legitimate predicate} $\leg : \alpha_i^n \implies \mathtt{row} \implies \mathtt{bool}$ takes $n$ parameters and returns a row predicate.
+A \emph{legitimate predicate} $\leg : \prod_i \alpha_i \implies \mathtt{row} \implies \mathtt{bool}$ takes $n$ parameters and returns a row predicate.
 For example, $\leg(x)(r_i) := \mathtt{int}(r_i(\text{``priors\_count''})) > x$ first take a parameter $x$ and then decides if priors count is larger than it.
 
-A \emph{calibration predicate} $\calib : \alpha_i^n \implies \mathtt{row} \implies \mathtt{bool}$, takes $n$ parameters and returns a row predicate.
+A \emph{calibration predicate} $\calib : \prod_i \alpha_i \implies \mathtt{row} \implies \mathtt{bool}$, takes $n$ parameters and returns a row predicate.
 For example, $\calib(u, l)(r_i) := l < \mathtt{int}(r_i(\text{``score''})) < u$ first takes two parameters $u,l$ as upper bound and lower bound, and then decides if $r_i$'s prediction score is between them.
@@
 \subsection{Definitions}
-We abstracted the idea of privileged groups and positive prediction as predicates to maximizes flexibility.
+We abstracted the idea of privileged groups and positive prediction as predicates to maximize flexibility.
 With the proper predicates, any model output can be used; even prose-like response of generative models can be included given adequate predicates.
 
-Given a datasets $\db$ for auditing use, and given csv prediction results or the prediction model $\model$ itself, we can calculate the fairness measures by modeling the statistical random variables mentioned in Section~\ref{sec:measures} with our predicates.
+Given a dataset $\db$ for auditing use, and given CSV prediction results or the prediction model $\model$ itself, we can calculate the fairness measures by modeling the statistical random variables mentioned in Section~\ref{sec:measures} with our predicates.
+We demonstrate this process in the following for a few selected measures; all the other measures can be modeled similarly.
 
-For a simple exmaple, for equal opportunity, recall its formal definition.
+For equal opportunity, recall its formal definition:
 \begin{align*}
@@
 \end{align*}
-To model the $Y$, $\hat{Y}$, and $S$, we we define the corresponding $\tru$, $\pos$, and $\priv$.
+To model the $Y$, $\hat{Y}$, and $S$, we define the corresponding $\tru$, $\pos$, and $\priv$.
 We have $Y = 1$ if and only if $\tru$ is true; $\hat{Y} = 1$ if and only if $\pos$ is true; $S = 1$ if and only if $\priv$ is true; and vice versa.
@@
 
-For positive balance, recall its formal definition.
+For positive balance, its formal definition is:
 \begin{align*}
@@
 
-For another exmaple, recall the formal definition of equal calibration.
+For equal calibration, recall its formal definition:
 \begin{align*}
@@
 
-where $\mathcal{M}$ is optional if the input contains csv prediction results.
+where $\mathcal{M}$ is optional if the input contains CSV prediction results.
 
-By calculating all of the available fairness measures, we provide the auditors a comprehensive perespective on the fairness performance of a model or dataset.
+By calculating all the available fairness measures, we provide the auditors a comprehensive perspective on the fairness performance of a model or dataset.
 
-We implemented the framework in Python and it is available as an public domain open-sourced Python package at \url{https://pypi.org/project/fairness-checker/}.
+We implemented the framework in Python, and it is available as a public domain open-sourced Python package at \url{https://pypi.org/project/fairness-checker/}.
 
@@
 Defendants taking the COMPAS test are given a questionnaire about topics ranging from family history to personal ideology.
-The questionnaire are then fed to the software system along with a number of parameters like the defendants's age, and then the system will assign a risk score to them from 1-10, with 10 being the highest risk.
+The questionnaire is then fed to the software system along with a number of parameters like the defendants' age, and then the system will assign a risk score to them from 1-10, with 10 being the highest risk.
 
@@
 They only counted misdemeanors and felonies as recidivism but not less serious crimes such as infractions.
-In the study, they've found that black defendants are disproportionately scored higher than they actually are and white defendants are disproportionately scored lower than they actually are.
+In the study, they've found that black defendants are disproportionately scored higher than they actually are, and white defendants are disproportionately scored lower than they actually are.
 
 We shall start applying our framework.
-There's generally no formal guide on how to set $\epsilon$, so we will take the 80\% rule's case and set it as $\epsilon = 0.2$.
+There is generally no formal guide on how to set $\epsilon$, so we will take the 80\% rule's case and set it as $\epsilon = 0.2$.
 For the $\epsilon$ of positive and negative balance, we shall use $\epsilon = 5$ as it is a reasonable choice.
-
-Although this dataset scenario falls in the csv-as-input case in our framework, and thus the model $\model$ input isn't needed, we also trained two simple makeshift models using the dataset itself for demonstration that simply outputs binary $\mathtt{bool}$ prediction, ``will recidivate'' and ``will not recidivate'', and $\mathtt{int}$ risk score, $1$ to $10$.
-The model was trained by splitting the original dataset into 3 partitions: 60\% of the data was used for training; 20\% was used for validation; and the rest 20\% was used for calculating fairness measures.
-The accuracy of the binary model was about 60\% and the accuracy of the score model was about 25\%.
 
@@
 
+For the score predicate, it is again a simple lookup.
+\[
+    \sco(r_i) := \mathtt{int}(r_i(\text{``decile\_score''}))
+\]
+
 For the legitimate predicate, we may want to look at defendants with priors. Here we shall set $args = (0)$.
@@
 
-For the calibration predicate, we may want to specifically look at risk scores between a specifically range. Here we shall set $args = (7,5)$.
+For the calibration predicate, we may want to look at risk scores between a specific range. Here we shall set $args = (7,5)$.
 \[
@@
 The results are compiled in Table~\ref{tab:results}.
-Since some of the fairness measures are equivalent, we write them in the same entry.
+Since some fairness measures are equivalent, we write them in the same entry.
 
@@
 On first blush, it is curious that it satisfies disparate impact but not demographic parity.
-However, if we return to the definition, we'd see that disparate impact is meant to be used when being marked positive is an advantaged thing, while here in the COMPAS exmaple, being marked positive is a disadvantaged thing.
+However, if we return to the definition, we'd see that disparate impact is meant to be used when being marked positive is an advantaged thing, while here in the COMPAS example, being marked positive is a disadvantaged thing.
 Thus, henceforth we will exclude disparate impact from our analysis.
@@
 From the results we can first notice that the overall accuracy eqaulity is low for both the Non-Asian and Non-Native American case.
-This can be explained by checking the original dataset which shows that there are only 32 and 18 rows, respectivly, in dataset of 7214 rows.
+This can be explained by checking the original dataset which shows that there are only 32 and 18 rows, respectively, in dataset of 7214 rows.
 Hence, the accuracy is naturally lower because of the small data size.
@@
 In another vein, the case of ``Male'' group and ``Misdemeanor'' group are both treated rather fairly.
-Furthermore, their overeal accuracy equality are both extremely good at the $< 0.1$ range.
+Furthermore, their overall accuracy equality are both excellent at the $< 0.1$ range.
 The predictive equality  of ``Male'' group even has a measure as low as $0.000004$.
+
+\subsection{Analysis with Model-as-input}
+Although this dataset scenario falls in the CSV-as-input case in our framework, we also trained a simple makeshift model using the dataset itself for demonstration.
+
+The model was trained by splitting the original dataset into 3 partitions: 60\% of the data was used for training; 20\% was used for validation; and the rest 20\% was used for calculating fairness measures.
+Validation showed that the accuracy was about 60\%.
+
+We only chose equalized odds of the case where privileged group is Non-African-American for illustrative purposes.
+The equalized odds of our model is 0.35 and 0.46, while that of COMPAS dataset is 0.23 and 0.22.
+So, our simple naive model is more unfair than COMPAS's model and could use some more fine-tuning.
 

### 2024-06-04 paper/paper.tex
One word change
Grammarly

@@
 With the rapid advancement of AI, there is a growing trend to integrate AI into decision-making processes.
-However, AI systems may exhibit biases that lead decision makers to draw unfair conclusions.
+However, AI systems may exhibit biases that lead decision-makers to draw unfair conclusions.
 Notably, the COMPAS system used in the American justice system to evaluate recidivism was found to favor racial majority groups; specifically, it violates a fairness standard called equalized odds.
 Various measures have been proposed to assess AI fairness.
-We present a framework for auditing AI fairness, involving third-party auditors and AI system provider, and we have created a tool to facilitate systematic examination of AI systems.
-Inclusion of independent and trusted auditors is pivotal for objectivity and accountability; third parties are often further necessary for their specialized expertise in relevant domains.
+We present a framework for auditing AI fairness, involving third-party auditors and AI system providers, and we have created a tool to facilitate systematic examination of AI systems.
+Including independent and trusted auditors is pivotal for objectivity and accountability; third parties are often further necessary for their specialized expertise in relevant domains.
 Auditors, equipped with our tool, can thoroughly review AI systems for bias and fairness violations.
@@
 Unlike traditional AI systems, we advocate a transparent white-box and statistics-based approach.
-It can be utilized by third party auditors, AI developers, or the general public for reference when judging the fairness criterion of AI systems.
+It can be utilized by third-party auditors, AI developers, or the general public for reference when judging the fairness criterion of AI systems.
 \end{abstract}
@@
 \section{Introduction}
-In recent years, the accelerating pace of artificial intelligence (AI) technologies have revolutionized numerous fields.
+The accelerating pace of artificial intelligence (AI) technologies has revolutionized numerous fields in recent years.
 In healthcare, AI-driven diagnostic systems streamline disease identification, while in finance, automated trading algorithms analyze market trends to execute optimal trades swiftly.
@@
 
-One of its most profound impact is how AI has transformed decision-making processes across sectors.
+One of its most profound impacts is how AI has transformed decision-making processes across sectors.
 By harnessing vast amounts of data and employing advanced algorithms, AI empowers organizations to make more informed and strategic decisions.
@@
 Factors such as inherent biases in original data sets or flaws in algorithm designs could contribute to bias within AI systems.
-When left unchecked, the ramifications of such biases extend far beyond mere inaccuracies; it can lead to devastating consequences, such as amplifying systemic injustices, perpetuating group discrimination, and exacerbating societal inequalities.
+When left unchecked, the ramifications of such biases extend far beyond mere inaccuracies; they can lead to devastating consequences, such as amplifying systemic injustices, perpetuating group discrimination, and exacerbating societal inequalities.
 
@@
 Fairness is fundamentally a subjective social construct, heavily influenced by cultural context and deeply rooted in historical inequalities.
-However, there have been development in research leveraging statistical metrics to quantify fairness that provide transparent and objective insights.
+However, there have been developments in research leveraging statistical metrics to quantify fairness that provide transparent and objective insights.
 
-By employing such mathematically rigorous methodologies, our research can offer a deeper look into the fairness of AI systems, enabling third parties to make more informed and robust judgment.
+By employing such mathematically rigorous methodologies, our research can offer a deeper look into the fairness of AI systems, enabling third parties to make more informed and robust judgments.
 For example, it would reveal that the COMPAS system violates a fairness measure called equalized odds.
@@
 
-With our tool, a third party such as ProPublica could clearly and easily demonstrates that COMPAS violates equalized odds according to their datasets.
+With our tool, a third party such as ProPublica could clearly and easily demonstrate that COMPAS violates equalized odds according to their datasets.
 Our platform empowers organizations to conduct thorough assessments of AI fairness and uncover discrepancies in predictive models.
@@
 
-Our tool is written in Python and is offered as a Python package. It supports common datasets format such as CSV. It is open-sourced and publicly available for downloads.
+Our tool is written in Python and is offered as a Python package. It supports common dataset formats such as CSV. It is open-sourced and publicly available for downloads.
 
@@
 \subsection{Preliminaries}
-For the following definition we will use $Y$ to denote the ground truth of an outcome; $\hat{Y}$ to denote the predicated result of an outcome; $Y = 1$ and $\hat{Y} = 1$ to denote them being accepted or positive.
+For the following definition, we will use $Y$ to denote the ground truth of an outcome; $\hat{Y}$ to denote the predicated result of an outcome; $Y = 1$ and $\hat{Y} = 1$ to denote them being accepted or positive.
 For example, let $Y$ be recidivism.
-Then $Y = 1$ means the case of an individual actually recidivating and $\hat{Y} = 0$ means the predication of an individual recidivating is negative.
-In addition, we use $V$ and $\hat{V}$ when the truth and the prediction isn't binary.
-For example, we denote COMPAS score by $\hat{V}$, which ranges from 1 to 10.
+Then $Y = 1$ means the case of an individual actually recidivating and $\hat{Y} = 0$ means the prediction of an individual recidivating is negative.
+In addition, we use $V$ and $\hat{V}$ when the truth and the prediction aren't binary.
+For example, we denote the COMPAS score by $\hat{V}$, which ranges from 1 to 10.
 
@@
 
-Legal cases involving disparate impact often refer to the ``80\% Rule'', advocated by the US Equal Employment Opportunity Commission \cite{eeoc1979}, where it requires the selection rate of a minority group is to be no less than 80\% of that of a majority group.
+Legal cases involving disparate impact often refer to the ``80\% Rule'', advocated by the US Equal Employment Opportunity Commission \cite{eeoc1979}, where it requires the selection rate of a minority group to be no less than 80\% of that of a majority group.
 Formally \cite{feldman2015certifying}, it is:
@@
 \subsection{Demographic Parity} %TODO: why cite this?
-Demographic parity \cite{calders2010three}, also known as statistical parity, is similar to disparate impact but, instead of ratio, difference is taken.
+Demographic parity \cite{calders2010three}, also known as statistical parity, is similar to disparate impact but, instead of ratio, the difference is taken.
 It is named so to suggest that each demographic group(such as race, gender, or age) should have equal representation or opportunity.
@@
 It considers the cases of true positive and false positive.
-It solves the downsides of a fully accurate classifier might be deemed unfair by measures that don't take ground truth into consideration.
+It solves the downsides of a fully accurate classifier that might be deemed unfair by measures that do not consider ground truth.
 For example, consider a group $A$ that is predicated to recidivate and they do in fact recidivate and a group $B$ that is predicated to recidivate but end up never recidivating.
@@
 \subsection{Conditional Use Accuracy Equality}
-Conditional use accuracy equality \cite{berk2021fairness} is similar to equalized odds, but instead of conditioning on the ground truth, it conditions on the prediction and calculated the probability of the ground truth.
+Conditional use accuracy equality \cite{berk2021fairness} is similar to equalized odds, but instead of conditioning on the ground truth, it conditions on the prediction and calculates the probability of the ground truth.
 It can be seen as checking the prediction accuracy across groups, thus the name.
-It further requires the measure on the case of positive predictive values is less than that of the negative predictive values.
+It further requires the measure in the case of positive predictive values to be less than that of the negative predictive values.
 Formally, it is:
@@
 For example, this could be conditioned on the highest COMPAS score $\hat{V} = 10$.
-Calibration is a concept of having fair score function \cite{fraenkel2020fairness}.
+Calibration is a concept of having a fair score function \cite{fraenkel2020fairness}.
 Formally, it is:
@@
 \label{sec:framework}
-Per the review by Pessach and Shmueli \cite{pessach2022review}, we designed an auditing framework for calculating the various fairness measure.
-We offer two versions of fairness checkers: one for when the prediction positive results are readily available in CSV input and one for when a model is provided.
-In most auditing cases the model version is preferred because CSV result can be easily fabricated.
+Per the review by Pessach and Shmueli \cite{pessach2022review}, we designed an auditing framework for calculating the various fairness measures.
+We offer two versions of fairness checkers: one for when the prediction results are readily available in CSV input and one for when a model is provided.
+In most auditing cases the model version is preferred because CSV results can be easily fabricated.
 
@@
 We abstracted the idea of privileged groups and positive prediction as predicates to maximize flexibility.
-With the proper predicates, any model output can be used; even prose-like response of generative models can be included given adequate predicates.
+With the proper predicates, any model output can be used; even prose-like responses of generative models can be included given adequate predicates.
 
@@
 
-This way, we can calculate each conditions programmatically by testing if the predicate is true on all rows and obtain the resulting measure.
+This way, we can calculate each condition programmatically by testing if the predicate is true on all rows and obtain the resulting measure.
 All the measures can be modeled in a similar vein.
@@
 
-Correctional Offender Management Profiling for Alternative Sanctions (COMPAS), developed by private company Northpointe (now Equivant), is a risk assessment software used in the American criminal justice system to evaluate the likelihood of a defendant reoffending.
+Correctional Offender Management Profiling for Alternative Sanctions (COMPAS), developed by the private company Northpointe (now Equivant), is a risk assessment software used in the American criminal justice system to evaluate the likelihood of a defendant reoffending.
 Defendants taking the COMPAS test are given a questionnaire about topics ranging from family history to personal ideology.
@@
 ProPublica is an American non-profit journalism organization focused on public interests.
-In 2016, they conduct an investigative report into the COMPAS system.
+In 2016, they conducted an investigative report into the COMPAS system.
 They obtained the 2013-2014 COMPAS score data of over 10,000 defendants in Florida.
@@
 
-For the calibration predicate, we may want to look at risk scores between a specific range. Here we shall set $args = (7,5)$.
+For the calibration predicate, we may want to look at risk scores within a specific range. Here we shall set $args = (7,5)$.
 \[
@@
 
-With these predicates defined, we can call the fairness measure functions and check if the measure hold or not.
+With these predicates defined, we can call the fairness measure functions and check if the measures hold or not.
 The results are compiled in Table~\ref{tab:results}.
@@
 
-We can then immediately tell from the failing demographic parity, mean difference, and conditional statistical parity that COMPAS prediction results was unfair against African-American, even if we only consider the ones with prior crimes.
+We can then immediately tell from the failing demographic parity, mean difference, and conditional statistical parity that COMPAS prediction results were unfair against African-American, even if we only consider the ones with prior crimes.
 
-From the low overall accuracy equality and both conditional use accuracy equality criteria we can tell that the accuracy is similar across African-Americans and non-African-Americans.
+From the low overall accuracy equality and both conditional use accuracy equality criteria, we can tell that the accuracy is similar across African-Americans and non-African-Americans.
 
@@
 It remains to be said if this is fair or not.
-An auditor could consult a domain expert for advice on how to set a $\epsilon$ and how is it justified.
+An auditor could consult a domain expert for advice on how to set a $\epsilon$ and how it is justified.
 
 \subsection{Fairness Analysis of Different Races}
-Setting the privilege predicate to different races we can have a more comprehensive look over the dataset.
+By setting the privilege predicate to different races we can have a more comprehensive look over the dataset.
 We've checked the case of privilege predicate being Non-African-American, Non-Asian, Non-Caucasian, Non-Hispanic, and Non-Native American.
@@
 
-From the results we can first notice that the overall accuracy eqaulity is low for both the Non-Asian and Non-Native American case.
-This can be explained by checking the original dataset which shows that there are only 32 and 18 rows, respectively, in dataset of 7214 rows.
+From the results, we can first notice that the overall accuracy equality is low for both the Non-Asian and Non-Native American cases.
+This can be explained by checking the original dataset which shows that there are only 32 and 18 rows, respectively, in a dataset of 7214 rows.
 Hence, the accuracy is naturally lower because of the small data size.
 
-Looking at the remaining three groups, Non-African-American, Non-Caucasian, and Non-Hispanic, their accuracy are relatively much better.
+Looking at the remaining three groups, Non-African-American, Non-Caucasian, and Non-Hispanic, their accuracies are relatively much better.
 
@@
 
-If we look closer we can see that in the age group ``Less than 25'', its predictive equality is noticeably worse, meaning in young people false positive rate is higher than older people; conversely, when we look at the age group ``Greater than 45'', both equal opportunity and predictive equality are worse, meaning for old people their treatment is even more unfair than young people.
+If we look closer we can see that in the age group ``Less than 25'', its predictive equality is noticeably worse, meaning in young people false positive rate is higher than in older people; conversely, when we look at the age group ``Greater than 45'', both equal opportunity and predictive equality are worse, meaning for old people their treatment is even more unfair than young people.
 Only the middle age group is treated fairly.
@@
 We only chose equalized odds of the case where privileged group is Non-African-American for illustrative purposes.
-The equalized odds of our model is 0.35 and 0.46, while that of COMPAS dataset is 0.23 and 0.22.
+The equalized odds of our model is 0.35 and 0.46, while that of the COMPAS dataset is 0.23 and 0.22.
 So, our simple naive model is more unfair than COMPAS's model and could use some more fine-tuning.
@@
 Addressing the challenges brought upon by the rapid advancements in AI technologies and its introduced bias necessitates a rigorous assessment of AI system fairness.
-While fairness is subjective and, it can be defined through various statistical measures.
+While fairness is subjective, it can be defined through various statistical measures.
 Our research contributes to this effort by proposing a comprehensive framework for auditing AI systems using multiple white-box fairness metrics, such as demographic parity, equalized odds, and overall accuracy equality.

### 2024-06-07 setup.py
Added fair string

@@ -5,7 +5,7 @@ with open("src/README.md", "r") as f:
 
 setup(
     name="fairness_checker",
-    version="0.0.3",
+    version="0.0.41",
     package_dir={'': 'src'},
     packages=find_packages(where="src"),
     description="Fairnes checker",

### 2024-06-07 src/fairness_checker/main.py

@@ -10,11 +10,12 @@ T = TypeVar('T')
 
 def print_stats(title, measure, ratio):
     print(title)
+    fair_string = "fair" if measure < ratio else "unfair"
     if measure >= 0.01:
-        print('fair: {0:.2f} < {1}'.format(measure, ratio))
+        print(f"{fair_string}: {measure:.2f} < {ratio}")
     else:
         precision = math.ceil(-math.log10(abs(measure) - abs(math.floor(measure))))
-        print('fair: {0:.{1}f} < {2}'.format(measure, precision, ratio))
+        print(f"{fair_string}: {measure:.{precision}f} < {ratio}")
 
 class fairness_model_checker:
     def __init__(self, raw_file: str):

### 2024-06-11 paper/paper.tex
todos

@@
 \newcommand{\pos}{\hat{P}}
+\newcommand{\posm}{\hat{P}_\mathcal{M}}
 \newcommand{\tru}{T}
@@
 \newcommand{\sco}{\hat{S}}
+\newcommand{\scom}{\hat{S}_\mathcal{M}}
 \newcommand{\calib}{C}
@@
 We present a framework for auditing AI fairness, involving third-party auditors and AI system providers, and we have created a tool to facilitate systematic examination of AI systems.
-Including independent and trusted auditors is pivotal for objectivity and accountability; third parties are often further necessary for their specialized expertise in relevant domains.
-Auditors, equipped with our tool, can thoroughly review AI systems for bias and fairness violations.
 The tool is open-sourced and publicly available.
@@
 
+%TODO: clarify the og defs and the role our tools plays
+% 1 fairness measures and compas eq odds
+% 2 our framework's role and it can do these calcs "and more"
+% 3 3rd party is important and with our tool...
+% 4 transparency
 By employing such mathematically rigorous methodologies, our research can offer a deeper look into the fairness of AI systems, enabling third parties to make more informed and robust judgments.
@@
 
+%TODO: fit this in
+Including independent and trusted auditors is pivotal for objectivity and accountability; third parties are often further necessary for their specialized expertise in relevant domains.
+Auditors, equipped with our tool, can thoroughly review AI systems for bias and fairness violations.
+
+%TODO:: rewrite following
 In addition to equalized odds, there are a number of fairness measures for AI systems.
@@
 
+%TODO: rewrite this. how is our tool transparent
 This functionality not only facilitates sound judgment but also fosters objectivity and transparency regarding AI systems, ultimately bolster the claim of fairness and justice in AI-enabled decision-making.
@@
 
-A positive outcome or prediction may be either beneficial or detrimental to the individual in different cases.
-Various measures were devised when considering different scenarios.
-For presentation, we consider $Y=1$ to be the detrimental case, like in the COMPAS prediction scenarios.
-
 \subsection{Disparate Impact}
@@
 
-\subsection{Demographic Parity} %TODO: why cite this?
-Demographic parity \cite{calders2010three}, also known as statistical parity, is similar to disparate impact but, instead of ratio, the difference is taken.
+\subsection{Demographic Parity}
+Demographic parity \cite{pessach2022review}, also known as statistical parity, is similar to disparate impact but, instead of ratio, the difference is taken.
 It is named so to suggest that each demographic group(such as race, gender, or age) should have equal representation or opportunity.
@@
 \subsection{Equal Opportunity}
-Equal opportunity \footnote{It was named so by considering the cases where positive outcomes benefit individuals, such as school admissions, hence the word opportunity.} \cite{hardt2016equality} is a relaxation of equalized odds by only considering the true positive case.
-Formally, it is:
+Equal opportunity \cite{hardt2016equality} is a relaxation of equalized odds by only considering the true positive case. Formally, it is:
 \begin{align*}
@@
 For example, $\pos(r_i) := \mathtt{int}(r_i(\text{``score''})) > 7$ means a row's prediction is positive if its score is greater than 7.
-A positive predicate of model results $\pos : \alpha \implies \mathtt{bool}$ takes a model's result and determines if it is positive.
+A positive predicate of model results $\posm : \alpha \implies \mathtt{bool}$ takes a model's result and determines if it is positive.
 In the simplest case, the model returns a $\mathtt{bool}$, and the positive predicate can just be the identity function.
@@
 A \emph{score predicate} of rows $\sco : \mathtt{row} \implies \alpha$ takes a row and gives its predicted score.
-Similar to positive predicate, a score predicate of model results $\sco : \alpha \implies \beta$ takes a model's result and gives its score.
+Similar to positive predicate, a score predicate of model results $\scom : \alpha \implies \beta$ takes a model's result and gives its score.
 
@@
 We have $Y = 1$ if and only if $\tru$ is true; $\hat{Y} = 1$ if and only if $\pos$ is true; $S = 1$ if and only if $\priv$ is true; and vice versa.
+%TODO: to calculate equal opp we call the function
 \begin{align*}
@@
 \end{align*}
-Here we model $Y$ and $S$ as above. As for $\hat{V} = v$, we abstracted $\hat{V}$ with $\calib$ and $v$ with $args$. We have $\hat{V} = v$ if and only if $\calib(args, ...)$ is true.
+Here we model $Y$ and $S$ as above. As for $\hat{V} = v$, we abstracted $\hat{V}$ with $\calib$ and $v$ with $args$. We have $\hat{V} = v$ if and only if $\calib(v, r_i)$ is true.
 \begin{align*}
-    \text{equal\_calibration} (\epsilon, \model, \priv, \calib, \tru, (args, ...))
+    \text{equal\_calibration} (\epsilon, \model, \priv, \calib, \tru, (v))
 \end{align*}
@@
 \]
-$\hat{Y}$ and $S$ are modeled as above. As for $L = l$, we abstracted it similarly to the case above by having $L = l$ if and only if $\leg(args, ...)$ is true.
+$\hat{Y}$ and $S$ are modeled as above. As for $L = l$, we abstracted it similarly to the case above by having $L = l$ if and only if $\leg(l)$ is true.
 \begin{align*}
-    \text{conditional\_statistical\_parity} (\epsilon, \model, \priv, \pos, \leg, (args, ...))
+    \text{conditional\_statistical\_parity} (\epsilon, \model, \priv, \pos, \leg, (l))
 \end{align*}
 
-This way, we can calculate each condition programmatically by testing if the predicate is true on all rows and obtain the resulting measure.
-All the measures can be modeled in a similar vein.
-So, in total, we have:
-
-\begin{align*}
-\text{disparate\_impact} & (\epsilon, \model, \priv, \pos) \\
-\text{demographic\_parity} & (\epsilon, \model, \priv, \pos) \\
-\text{conditional\_statistical\_parity} & (\epsilon, \model, \priv, \pos, \leg, (args, ...)) \\
-\text{overall\_accuracy\_eqaulity} & (\epsilon, \model, \priv, \pos, \tru) \\
-\text{mean\_difference} & (\epsilon, \model, \priv, \pos) \\
-\text{equalized\_odds} & (\epsilon, \model, \priv, \pos, \tru) \\
-\text{equal\_opportunity} & (\epsilon, \model, \priv, \pos, \tru) \\
-\text{predictive\_equality} & (\epsilon, \model, \priv, \pos, \tru) \\
-\text{conditional\_use\_accuracy\_equality} & (\epsilon, \model, \priv, \pos, \tru) \\
-\text{predictive\_parity} & (\epsilon, \model, \priv, \pos, \tru) \\
-\text{equal\_calibration} & (\epsilon, \model, \priv, \calib, \tru, (args, ...)) \\
-\text{positive\_balance} & (\epsilon, \model, \priv, \sco, \tru) \\
-\text{nagative\_balance} & (\epsilon, \model, \priv, \sco, \tru) \\
-\end{align*}
-
-where $\mathcal{M}$ is optional if the input contains CSV prediction results.
+This way, we can calculate each condition programmatically by testing if the predicate is true on all rows and obtain the resulting measure. All the measures can be modeled in a similar vein.
 
@@
 They only counted misdemeanors and felonies as recidivism but not less serious crimes such as infractions.
-In the study, they've found that black defendants are disproportionately scored higher than they actually are, and white defendants are disproportionately scored lower than they actually are.
+In the study, they have found that black defendants are disproportionately scored higher than they actually are, and white defendants are disproportionately scored lower than they actually are.
 
-We shall start applying our framework.
-There is generally no formal guide on how to set $\epsilon$, so we will take the 80\% rule's case and set it as $\epsilon = 0.2$.
-For the $\epsilon$ of positive and negative balance, we shall use $\epsilon = 5$ as it is a reasonable choice.
+We shall start applying our framework. There is generally no formal guide on how to set $\epsilon$, so we will take the 80\% rule's case and set it as $\epsilon = 0.2$.
 
-We will set the privileged predicate be to ``non-African-American'' so that the unprivileged group will be African-American to test if the African-American race is discriminated.
+We will set the unprivileged predicate be to ``African-American'' so that the unprivileged group will be African-American to test if the African-American race is discriminated.
 \[
@@
 \]
-Alternatively, since we know the corresponding scores of the categories, we can define it using the scores themselves, too.
-\[
-    \pos(r_i) := 5 \leq \mathtt{int}(r_i(\text{``decile\_score''})) \leq 10
-\]
-On the other hand, when we're taking prediction input from our makeshift model, we simply return its result because it already gives a binary $\mathtt{bool}$ result: $\pos(r_i) := \model(r_i)$
+On the other hand, when we're taking prediction input from our makeshift model, we simply return its result because it already gives a binary $\mathtt{bool}$ result: $\posm(r_i) := \model(r_i)$
 
@@
 
-For the legitimate predicate, we may want to look at defendants with priors. Here we shall set $args = (0)$.
+For the legitimate predicate, we may want to look at defendants with priors.
 \[
@@
 
-For the calibration predicate, we may want to look at risk scores within a specific range. Here we shall set $args = (7,5)$.
+For the calibration predicate, we may want to look at risk scores within a specific range.
 \[
@@
     \end{tabular}
-    \caption{Fairness measures of privileged group being Non-African-American.}
+    \caption{Fairness measures of unprivileged group being African-American.}
     \label{tab:results}
@@
 \subsection{Fairness Analysis of African-American group}
+%TODO: explain why other measures still work because they're taking the difference
 On first blush, it is curious that it satisfies disparate impact but not demographic parity.
-However, if we return to the definition, we'd see that disparate impact is meant to be used when being marked positive is an advantaged thing, while here in the COMPAS example, being marked positive is a disadvantaged thing.
+However, if we return to the definition, we would see that disparate impact is meant to be used when being marked positive is an advantaged thing, while here in the COMPAS example, being marked positive is a disadvantaged thing.
 Thus, henceforth we will exclude disparate impact from our analysis.
@@
 
-From the low equal calibration we can tell that if we only consider the medium risk score, African-Americans aren't treated fairly.
+From the low equal calibration we can tell that if we only consider the medium risk score, African-Americans are treated fairly.
 
@@
 By setting the privilege predicate to different races we can have a more comprehensive look over the dataset.
-We've checked the case of privilege predicate being Non-African-American, Non-Asian, Non-Caucasian, Non-Hispanic, and Non-Native American.
+We have checked the case of unprivilege predicate being African-American, Asian, Caucasian, Hispanic, and Native American.
 The results are shown in Figure~\ref{fig:compas}.
 
-From the results, we can first notice that the overall accuracy equality is low for both the Non-Asian and Non-Native American cases.
+From the results, we can first notice that the overall accuracy equality is low for both the Asian and Native American cases.
 This can be explained by checking the original dataset which shows that there are only 32 and 18 rows, respectively, in a dataset of 7214 rows.
@@
 
-Looking at the remaining three groups, Non-African-American, Non-Caucasian, and Non-Hispanic, their accuracies are relatively much better.
+Looking at the remaining three groups, African-American, Caucasian, and Hispanic, their accuracies are relatively much better.
 
-As for fairness measures, we can immediately see from that all measures are $\epsilon < 0.2$ for Non-Caucasian and Non-Hispanic that African-American are indeed treated unfairly in a broad sense.
+As for fairness measures, we can immediately see from that all measures are $\epsilon < 0.2$ for Caucasian and Hispanic that African-American are indeed treated unfairly in a broad sense.
 
@@
 
-We only chose equalized odds of the case where privileged group is Non-African-American for illustrative purposes.
+We only chose equalized odds of the case where unprivileged group is African-American for illustrative purposes.
 The equalized odds of our model is 0.35 and 0.46, while that of the COMPAS dataset is 0.23 and 0.22.
@@
         \includegraphics[width=\linewidth]{Non-African-American}
-        \subcaption{Privileged Group: Non-African-American}\label{fig:1}
+        \subcaption{Privileged Group: African-American}\label{fig:1}
     \end{minipage}
@@
         \includegraphics[width=\linewidth]{Non-Caucasian}
-        \subcaption{Privileged Group: Non-Caucasian}\label{fig:2}
+        \subcaption{Privileged Group: Caucasian}\label{fig:2}
     \end{minipage}
@@
         \includegraphics[width=\linewidth]{Non-Asian}
-        \subcaption{Privileged Group: Non-Asian}\label{fig:2}
+        \subcaption{Privileged Group: Asian}\label{fig:2}
     \end{minipage}
@@
         \includegraphics[width=\linewidth]{Non-Native American}
-        \subcaption{Privileged Group: Non-Native American}\label{fig:2}
+        \subcaption{Privileged Group: Native American}\label{fig:2}
     \end{minipage}
@@
         \includegraphics[width=\linewidth]{Non-Hispanic}
-        \subcaption{Privileged Group: Non-Hispanic}\label{fig:2}
+        \subcaption{Privileged Group: Hispanic}\label{fig:2}
     \end{minipage}
@@
         \includegraphics[width=\linewidth]{Non-Less than 25}
-        \subcaption{Privileged Group: Non-Less than 25}\label{fig:1}
+        \subcaption{Privileged Group: Greater than 25}\label{fig:1}
     \end{minipage}
@@
         \includegraphics[width=\linewidth]{Non-25 - 45}
-        \subcaption{Privileged Group: Non-25 - 45}\label{fig:2}
+        \subcaption{Privileged Group: Less than 25 or Greater than 45}\label{fig:2}
     \end{minipage}
@@
         \includegraphics[width=\linewidth]{Non-Greater than 45}
-        \subcaption{Privileged Group: Non-Greater than 45}\label{fig:2}
+        \subcaption{Privileged Group: Less than 45}\label{fig:2}
     \end{minipage}

### 2024-06-12 paper/paper.tex
Major revision

@@
 
-%TODO: clarify the og defs and the role our tools plays
-% 1 fairness measures and compas eq odds
-% 2 our framework's role and it can do these calcs "and more"
-% 3 3rd party is important and with our tool...
-% 4 transparency
-By employing such mathematically rigorous methodologies, our research can offer a deeper look into the fairness of AI systems, enabling third parties to make more informed and robust judgments.
-For example, it would reveal that the COMPAS system violates a fairness measure called equalized odds.
+These statistical metrics offer a systematic approach to evaluating fairness across various dimensions of AI systems.
+Among these metrics are disparate impact and demographic parity, to name a few.
+In particular, the findings of the COMPAS report revealed that it violates a critical metric known as equalized odds.
 
-%TODO: fit this in
-Including independent and trusted auditors is pivotal for objectivity and accountability; third parties are often further necessary for their specialized expertise in relevant domains.
+The function of our framework is to make comprehensive reviewing of statistical metrics accessible to third party auditors.
+Central to this role is the precise definition and flexible abstraction of the metrics.
+For example, our framework would reveal that the COMPAS dataset not only violates equalized odds but also conditional statistical parity and mean difference.
+
+Including independent and trusted auditors is pivotal for objectivity and accountability.
+Moreover, the involvement of third parties is often indispensable due to their specialized expertise in relevant domains.
 Auditors, equipped with our tool, can thoroughly review AI systems for bias and fairness violations.
 
-%TODO:: rewrite following
-In addition to equalized odds, there are a number of fairness measures for AI systems.
-Our framework aims to help third parties apply these metrics and navigate complex datasets with ease, pinpointing instances of bias and discrimination.
+Our tool builds upon transparent definitions widely accepted within the scientific community.
+By providing an abstraction layer atop these definitions, we enhance their flexibility and transparency, enabling users to adapt them to diverse contexts with ease and reveal them to the public.
+Moreover, we prioritize transparency throughout our tool's development process, offering access to the source code for scrutiny, thereby ensuring transparency from conception to implementation.
 
-With our tool, a third party such as ProPublica could clearly and easily demonstrate that COMPAS violates equalized odds according to their datasets.
-Our platform empowers organizations to conduct thorough assessments of AI fairness and uncover discrepancies in predictive models.
-
-%TODO: rewrite this. how is our tool transparent
-This functionality not only facilitates sound judgment but also fosters objectivity and transparency regarding AI systems, ultimately bolster the claim of fairness and justice in AI-enabled decision-making.
-
-Our tool is written in Python and is offered as a Python package. It supports common dataset formats such as CSV. It is open-sourced and publicly available for downloads.
+Our open-sourced tool is written in Python and is offered as a Python package.
+It supports common dataset formats such as CSV.
+The public domain source code is available at \url{https://github.com/RexYuan/Shu}.
 
@@
 We have $Y = 1$ if and only if $\tru$ is true; $\hat{Y} = 1$ if and only if $\pos$ is true; $S = 1$ if and only if $\priv$ is true; and vice versa.
-%TODO: to calculate equal opp we call the function
+This way, we can calculate equal opportunity as a function:
 \begin{align*}
@@
 \end{align*}
-We model $Y$ and $S$ as the previous example. As for $\hat{V}$ we have it equal to the output of $\sco$, and now we can calculate the measure.
+We model $Y$ and $S$ as the previous example. As for $\hat{V}$ we have it equal to the output of $\sco$, and now we can calculate the measure. We calculate positive balance by calling the function:
 \begin{align*}
@@
 \end{align*}
-Here we model $Y$ and $S$ as above. As for $\hat{V} = v$, we abstracted $\hat{V}$ with $\calib$ and $v$ with $args$. We have $\hat{V} = v$ if and only if $\calib(v, r_i)$ is true.
+Here we model $Y$ and $S$ as above.
+As for $\hat{V} = v$, we abstracted $\hat{V}$ with $\calib$ and $v$ with $args$.
+We have $\hat{V} = v$ if and only if $\calib(v)(r_i)$ is true.
+We calculate equal calibration by calling the function:
 \begin{align*}
@@
 \]
-$\hat{Y}$ and $S$ are modeled as above. As for $L = l$, we abstracted it similarly to the case above by having $L = l$ if and only if $\leg(l)$ is true.
+$\hat{Y}$ and $S$ are modeled as above.
+As for $L = l$, we abstracted it similarly to the case above by having $L = l$ if and only if $\leg(l)(r_i)$ is true.
+We calculate conditional statistical parity by calling the function:
 \begin{align*}
@@
 \subsection{Fairness Analysis of African-American group}
-%TODO: explain why other measures still work because they're taking the difference
 On first blush, it is curious that it satisfies disparate impact but not demographic parity.
 However, if we return to the definition, we would see that disparate impact is meant to be used when being marked positive is an advantaged thing, while here in the COMPAS example, being marked positive is a disadvantaged thing.
-Thus, henceforth we will exclude disparate impact from our analysis.
+
+It deceives the eye due to its definition being a ratio that lacks symmetry upon interchange of its components.
+Henceforth we will exclude disparate impact from our analysis of COMPAS.
+On the contrary, the other measures remain unaffected since their definitions are the absolute values of a difference.
 

### 2024-06-13 setup.py
Added command line CLI interface

@@ -5,7 +5,7 @@ with open("src/README.md", "r") as f:
 
 setup(
     name="fairness_checker",
-    version="0.0.41",
+    version="0.0.5",
     package_dir={'': 'src'},
     packages=find_packages(where="src"),
     description="Fairnes checker",
@@ -16,4 +16,9 @@ setup(
     author_email="hello@rexyuan.com",
     license="Unlicense",
     python_requires=">=3.6",
+    entry_points={
+        'console_scripts': [
+            'fairness_checker=fairness_checker.__main__:main'
+        ]
+    },
 )

### 2024-06-13 src/fairness_checker/main.py

@@ -296,9 +296,9 @@ class fairness_model_checker:
                           ratio: float,
                           model: object,
                           privileged_predicate: Callable[[csv_row], bool],
+                          truth_predicate: Callable[[csv_row], bool],
                           calib_predicate_h: Callable[..., Callable[[csv_row], bool]],
-                          calib_arg: Tuple[Any, ...],
-                          truth_predicate: Callable[[csv_row], bool]) -> bool:
+                          calib_arg: Tuple[Any, ...]) -> bool:
         """
         Evaluates the equal calibration of the model's predictions between privileged and unprivileged groups.
 
@@ -521,14 +521,14 @@ class fairness_model_checker:
 
         return measure < ratio
 
-    def nagative_balance(self,
+    def negative_balance(self,
                          ratio: float,
                          model: object,
                          privileged_predicate: Callable[[csv_row], bool],
                          score_predicate: Callable[[csv_row], bool],
                          truth_predicate: Callable[[csv_row], bool]) -> bool:
         """
-        Evaluates the nagative balance of the model's predictions between privileged and unprivileged groups.
+        Evaluates the negative balance of the model's predictions between privileged and unprivileged groups.
 
         Parameters:
         ratio (float): The fairness threshold ratio.
@@ -820,9 +820,9 @@ class fairness_csv_checker:
     def equal_calibration(self,
                           ratio: float,
                           privileged_predicate: Callable[[csv_row], bool],
+                          truth_predicate: Callable[[csv_row], bool],
                           calib_predicate_h: Callable[..., Callable[[csv_row], bool]],
-                          calib_arg: Tuple[Any, ...],
-                          truth_predicate: Callable[[csv_row], bool]) -> bool:
+                          calib_arg: Tuple[Any, ...]) -> bool:
         """
         Evaluates the equal calibration of the model's predictions between privileged and unprivileged groups.
 
@@ -988,13 +988,13 @@ class fairness_csv_checker:
 
         return measure < ratio
 
-    def nagative_balance(self,
+    def negative_balance(self,
                          ratio: float,
                          privileged_predicate: Callable[[csv_row], bool],
                          score_predicate: Callable[[csv_row], bool],
                          truth_predicate: Callable[[csv_row], bool]) -> bool:
         """
-        Evaluates the nagative balance of the model's predictions between privileged and unprivileged groups.
+        Evaluates the negative balance of the model's predictions between privileged and unprivileged groups.
 
         Parameters:
         ratio (float): The fairness threshold ratio.

### 2024-06-13 src/test/test_csv.py

@@ -13,10 +13,10 @@ c.equalized_odds(0.2, lambda row:                    row['sex'] == sex, lambda r
 c.equal_opportunity(0.2, lambda row:                 row['sex'] == sex, lambda row: row['score_text'] in {'Medium', 'High'}, lambda row: row['is_recid'] == '1')
 c.accuracy_eqaulity(0.2, lambda row:                 row['sex'] == sex, lambda row: row['score_text'] in {'Medium', 'High'}, lambda row: row['is_recid'] == '1')
 c.predictive_parity(0.2, lambda row:                 row['sex'] == sex, lambda row: row['score_text'] in {'Medium', 'High'}, lambda row: row['is_recid'] == '1')
-c.equal_calibration(0.2, lambda row:                 row['sex'] == sex, lambda u,l: (lambda row: l <= int(row['decile_score']) and int(row['decile_score']) <= u), (7, 5), lambda row: row['is_recid'] == '1')
+c.equal_calibration(0.2, lambda row:                 row['sex'] == sex, lambda row: row['is_recid'] == '1', lambda u,l: (lambda row: l <= int(row['decile_score']) and int(row['decile_score']) <= u), (7, 5))
 c.conditional_statistical_parity(0.2, lambda row:    row['sex'] == sex, lambda row: row['score_text'] in {'Medium', 'High'}, lambda x: (lambda row: int(row['priors_count']) > x), (0,))
 c.predictive_equality(0.2, lambda row:               row['sex'] == sex, lambda row: row['score_text'] in {'Medium', 'High'}, lambda row: row['is_recid'] == '1')
 c.conditional_use_accuracy_equality(0.2, lambda row: row['sex'] == sex, lambda row: row['score_text'] in {'Medium', 'High'}, lambda row: row['is_recid'] == '1')
 c.positive_balance(5, lambda row:                    row['sex'] == sex, lambda row: int(row['decile_score']), lambda row: row['is_recid'] == '1')
-c.nagative_balance(5, lambda row:                    row['sex'] == sex, lambda row: int(row['decile_score']), lambda row: row['is_recid'] == '1')
+c.negative_balance(5, lambda row:                    row['sex'] == sex, lambda row: int(row['decile_score']), lambda row: row['is_recid'] == '1')
 c.mean_difference(0.2, lambda row:                   row['sex'] == sex, lambda row: row['score_text'] in {'Medium', 'High'})

### 2024-06-13 src/test/test_model.py

@@ -119,10 +119,10 @@ c.equalized_odds(0.2, trained, lambda row: row['race'] != race, lambda Y: Y == 1
 c.equal_opportunity(0.2, trained, lambda row: row['race'] != race, lambda Y: Y == 1, lambda row: row['two_year_recid'] == '1')
 c.accuracy_eqaulity(0.2, trained, lambda row: row['race'] != race, lambda Y: Y == 1, lambda row: row['is_recid'] == '1')
 c.predictive_parity(0.2, trained, lambda row: row['race'] != race, lambda Y: Y == 1, lambda row: row['is_recid'] == '1')
-c.equal_calibration(0.2, trained_tree, lambda row: row['race'] != race, lambda u,l: (lambda Y: l <= int(Y) and int(Y) <= u), (7, 5), lambda row: row['is_recid'] == '1')
+c.equal_calibration(0.2, trained_tree, lambda row: row['race'] != race, lambda row: row['is_recid'] == '1', lambda u,l: (lambda Y: l <= int(Y) and int(Y) <= u), (7, 5))
 c.conditional_statistical_parity(0.2, trained, lambda row: row['race'] != race, lambda Y: Y == 1, lambda x: (lambda row: int(row['priors_count']) > x), (0,))
 c.predictive_equality(0.2, trained, lambda row: row['race'] != race, lambda Y: Y == 1, lambda row: row['is_recid'] == '1')
 c.conditional_use_accuracy_equality(0.2, trained, lambda row: row['race'] != race, lambda Y: Y == 1, lambda row: row['is_recid'] == '1')
 c.positive_balance(5, trained_tree, lambda row: row['race'] != race, lambda Y: int(Y), lambda row: row['is_recid'] == '1')
-c.nagative_balance(5, trained_tree, lambda row: row['race'] != race, lambda Y: int(Y), lambda row: row['is_recid'] == '1')
+c.negative_balance(5, trained_tree, lambda row: row['race'] != race, lambda Y: int(Y), lambda row: row['is_recid'] == '1')
 c.mean_difference(0.2, trained, lambda row: row['race'] != race, lambda Y: Y == 1)

### 2024-06-13 src/test/test_predicates1.py

@@ -1,0 +1,8 @@
+def privileged_predicate(row):
+    return row['sex'] == 'Male'
+
+def score_predicate(row):
+    return int(row['decile_score'])
+
+def truth_predicate(row):
+    return row['is_recid'] == '1'

### 2024-06-13 src/test/test_predicates2.py

@@ -1,0 +1,12 @@
+def privileged_predicate(row):
+    return row['sex'] == 'Male'
+
+def truth_predicate(row):
+    return row['is_recid'] == '1'
+
+def calib_predicate_h(u, l):
+    def tmp(row):
+        return l <= int(row['decile_score']) and int(row['decile_score']) <= u
+    return tmp
+
+calib_arg = (7, 5)

### 2024-06-13 setup.py
bump version

@@ -5,7 +5,7 @@ with open("src/README.md", "r") as f:
 
 setup(
     name="fairness_checker",
-    version="0.0.5",
+    version="0.1.0",
     package_dir={'': 'src'},
     packages=find_packages(where="src"),
     description="Fairnes checker",

### 2024-06-18 paper/paper.tex
minor revision
Final version

@@
 \title{Ensuring Fairness with Transparent Auditing of Quantitative Bias in AI Systems\\
-\thanks{Identify applicable funding agency here. If none, delete this.}
-}
-
-\author{\IEEEauthorblockN{1\textsuperscript{st} Given Name Surname}
-\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-\textit{name of organization (of Aff.)}\\
-City, Country \\
-email address or ORCID}
-\and
-\IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
-\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-\textit{name of organization (of Aff.)}\\
-City, Country \\
-email address or ORCID}
+\thanks{This project is partially supported by NSTC 112-2634-F-002-004 from National Science and Technology Council and AS-GCS-113-H04 from Academia Sinica.}
 }
@@
 
-% \author{\IEEEauthorblockN{1\textsuperscript{st} Chih-Cheng Rex Yuan}
-% \IEEEauthorblockA{\textit{Institute of Information Science} \\
-% \textit{Academia Sinica}\\
-% Taipei, Taiwan \\
-% hello@rexyuan.com}
-% \and
-% \IEEEauthorblockN{2\textsuperscript{nd} Bow-Yaw Wang}
-% \IEEEauthorblockA{\textit{Institute of Information Science} \\
-% \textit{Academia Sinica}\\
-% Taipei, Taiwan \\
-% bywang@iis.sinica.edu.tw}
-% }
+\author{\IEEEauthorblockN{Chih-Cheng Rex Yuan}
+\IEEEauthorblockA{\textit{Institute of Information Science} \\
+\textit{Academia Sinica}\\
+Taipei, Taiwan \\
+hello@rexyuan.com}
+\and
+\IEEEauthorblockN{Bow-Yaw Wang}
+\IEEEauthorblockA{\textit{Institute of Information Science} \\
+\textit{Academia Sinica}\\
+Taipei, Taiwan \\
+bywang@iis.sinica.edu.tw}
+}
 
@@
 
-AI system, however, may be biased.
+AI systems, however, may be biased.
 Factors such as inherent biases in original data sets or flaws in algorithm designs could contribute to bias within AI systems.
@@
 Our tool builds upon transparent definitions widely accepted within the scientific community.
-By providing an abstraction layer atop these definitions, we enhance their flexibility and transparency, enabling users to adapt them to diverse contexts with ease and reveal them to the public.
+By providing an abstraction layer atop these definitions, we enhance their flexibility and transparency, which enables users to adapt them to diverse contexts with ease and reveal them to the public.
 Moreover, we prioritize transparency throughout our tool's development process, offering access to the source code for scrutiny, thereby ensuring transparency from conception to implementation.
@@
 It supports common dataset formats such as CSV.
-The public domain source code is available at \url{https://github.com/RexYuan/Shu}.
+The package is available at \url{https://pypi.org/project/fairness-checker/}.
 
@@
 Then $Y = 1$ means the case of an individual actually recidivating and $\hat{Y} = 0$ means the prediction of an individual recidivating is negative.
-In addition, we use $V$ and $\hat{V}$ when the truth and the prediction aren't binary.
+In addition, we use $V$ and $\hat{V}$ when the truth and the prediction are not binary.
 For example, we denote the COMPAS score by $\hat{V}$, which ranges from 1 to 10.
@@
 
-A \emph{score predicate} of rows $\sco : \mathtt{row} \implies \alpha$ takes a row and gives its predicted score.
+A \emph{score predicate} of rows $\sco : \mathtt{row} \implies \beta$ takes a row and gives its predicted score of some type $\beta$.
 Similar to positive predicate, a score predicate of model results $\scom : \alpha \implies \beta$ takes a model's result and gives its score.
@@
 A \emph{legitimate predicate} $\leg : \prod_i \alpha_i \implies \mathtt{row} \implies \mathtt{bool}$ takes $n$ parameters and returns a row predicate.
-For example, $\leg(x)(r_i) := \mathtt{int}(r_i(\text{``priors\_count''})) > x$ first take a parameter $x$ and then decides if priors count is larger than it.
+For example, $\leg(x)(r_i) := \mathtt{int}(r_i(\text{``priors\_count''})) > x$ first takes a parameter $x$ and then decides if priors count is larger than it.
 
@@
 We abstracted the idea of privileged groups and positive prediction as predicates to maximize flexibility.
-With the proper predicates, any model output can be used; even prose-like responses of generative models can be included given adequate predicates.
+With proper predicates, any model output can be used; even prose-like responses of generative models can be included given adequate predicates.
 
@@
 \end{align*}
-To model the $Y$, $\hat{Y}$, and $S$, we define the corresponding $\tru$, $\pos$, and $\priv$.
+To model $Y$, $\hat{Y}$, and $S$, we define the corresponding $\tru$, $\pos$, and $\priv$.
 We have $Y = 1$ if and only if $\tru$ is true; $\hat{Y} = 1$ if and only if $\pos$ is true; $S = 1$ if and only if $\priv$ is true; and vice versa.
@@
 
-We will set the unprivileged predicate be to ``African-American'' so that the unprivileged group will be African-American to test if the African-American race is discriminated.
+We will set the unprivileged predicate to be ``African-American'' so that the unprivileged group will be African-American. Then we test if the African-American race is discriminated.
 \[
@@
 \]
-On the other hand, when we're taking prediction input from our makeshift model, we simply return its result because it already gives a binary $\mathtt{bool}$ result: $\posm(r_i) := \model(r_i)$
 
@@
         \centering
-        \includegraphics[width=\linewidth]{Non-African-American}
-        \subcaption{Privileged Group: African-American}\label{fig:1}
+        \includegraphics[width=\linewidth]{African-American}
+        \subcaption{Unprivileged Group: African-American}\label{fig:1}
     \end{minipage}
@@
         \centering
-        \includegraphics[width=\linewidth]{Non-Caucasian}
-        \subcaption{Privileged Group: Caucasian}\label{fig:2}
+        \includegraphics[width=\linewidth]{Caucasian}
+        \subcaption{Unprivileged Group: Caucasian}\label{fig:2}
     \end{minipage}
@@
         \centering
-        \includegraphics[width=\linewidth]{Non-Asian}
-        \subcaption{Privileged Group: Asian}\label{fig:2}
+        \includegraphics[width=\linewidth]{Asian}
+        \subcaption{Unprivileged Group: Asian}\label{fig:2}
     \end{minipage}
@@
         \centering
-        \includegraphics[width=\linewidth]{Non-Native American}
-        \subcaption{Privileged Group: Native American}\label{fig:2}
+        \includegraphics[width=\linewidth]{Native American}
+        \subcaption{Unprivileged Group: Native American}\label{fig:2}
     \end{minipage}
@@
         \centering
-        \includegraphics[width=\linewidth]{Non-Hispanic}
-        \subcaption{Privileged Group: Hispanic}\label{fig:2}
+        \includegraphics[width=\linewidth]{Hispanic}
+        \subcaption{Unprivileged Group: Hispanic}\label{fig:2}
     \end{minipage}
@@
         \centering
-        \includegraphics[width=\linewidth]{Non-Less than 25}
-        \subcaption{Privileged Group: Greater than 25}\label{fig:1}
+        \includegraphics[width=\linewidth]{Less than 25}
+        \subcaption{Unprivileged Group: Less than 25}\label{fig:1}
     \end{minipage}
@@
         \centering
-        \includegraphics[width=\linewidth]{Non-25 - 45}
-        \subcaption{Privileged Group: Less than 25 or Greater than 45}\label{fig:2}
+        \includegraphics[width=\linewidth]{25 - 45}
+        \subcaption{Unprivileged Group: 25 - 45}\label{fig:2}
     \end{minipage}
@@
         \centering
-        \includegraphics[width=\linewidth]{Non-Greater than 45}
-        \subcaption{Privileged Group: Less than 45}\label{fig:2}
+        \includegraphics[width=\linewidth]{Greater than 45}
+        \subcaption{Unprivileged Group: Greater than 45}\label{fig:2}
     \end{minipage}

### 2024-07-04 paper/paper.tex
Update funding source

@@
 \title{Ensuring Fairness with Transparent Auditing of Quantitative Bias in AI Systems\\
-\thanks{This project is partially supported by NSTC 112-2634-F-002-004 from National Science and Technology Council and AS-GCS-113-H04 from Academia Sinica.}
+\thanks{This project is partially supported by MOST111-2221-E-001-010-MY2 and 112-2634-F-002-004 from National Science and Technology Council and AS-GCS-113-H04 from Academia Sinica.}
 }

### 2024-08-16 slides/2024_8_20.tex
Added slides

\documentclass{beamer}

\usetheme{CambridgeUS}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

\usepackage{CJKutf8}
\usepackage{datetime}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{subcaption}
\usepackage{xcolor}
\usepackage{ulem}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{cite}
\usepackage{multirow}
\usepackage{makecell}
\usepackage{url}
\usepackage{placeins}

\graphicspath{{images/}}
\DeclareGraphicsExtensions{.pdf}

\newdate{date}{20}{08}{2024}
\date{\displaydate{date}}
\title[Fairness]{Ensuring Fairness with Transparent Auditing of
Quantitative Bias in AI Systems}
\author[Rex]{
    \begin{CJK}{UTF8}{bsmi}袁至誠\end{CJK}\newline
    Chih-cheng Rex Yuan\newline
    \href{https://rexyuan.com/}{rexyuan.com}
    }
\institute[IIS,AS]{Institute of Information Science, Academia Sinica}

\DeclarePairedDelimiter{\set}{\{}{\}}
\DeclarePairedDelimiter{\tuple}{(}{)}
\DeclarePairedDelimiter{\abs}{\lvert}{\rvert}

\let\oldleq\leq
\renewcommand{\leq}[1][]{\oldleq_{#1}}
\renewcommand{\implies}{\rightarrow}

\newcommand{\bye}[1]{}

\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\sred}[1]{\textcolor{red}{\sout{#1}}}

\begin{document}

\begin{frame}
\titlepage
\end{frame}

\begin{frame}
    \frametitle{Bias}
    \begin{itemize}
        \item AI is widely used in decision making:
        \begin{itemize}
            \item School admission.
            \item Loan approval.
            \item Hiring.
            \item Policing.
            \item Censorship.
            \item etc..
        \end{itemize}
        \item Decision making by AI may be biased.
        \item Bias can come from several sources:
        \begin{itemize}
            \item Biased data. ML is deisgned to replicate this.
            \item Missing data. The datasets might not be representative.
            \item Biased algorithms. The objective functions might introduce bias.
            \item Sensitive attributes: Age, Gender, ..., etc..
        \end{itemize}
    \end{itemize}
\end{frame}

\begin{frame}
    \frametitle{Protected Attributes}
    What are the protected(sensitive) attributes?
    \begin{table}
        \begin{tabular}{|c|c|c|c|c|}
            \hline
            \red{Age} & \red{Gender} & Occupation & \red{Income} & Education \\
            \hline
            \red{28} & \red{M} & Engineering & \red{\$80,000} & Master \\
            \red{28} & \red{F} & Engineering & \red{\$65,000} & Master \\
            \red{45} & \red{M} & Medicine    & \red{\$100,000} & Doctorate \\
            \red{40} & \red{F} & Legal       & \red{\$150,000} & Law Degree \\
            \red{32} & \red{M} & Education   & \red{\$55,000} & Bachelor \\
            \hline
        \end{tabular}
        \caption{Example Dataset}
    \end{table}
\end{frame}

\begin{frame}
    \frametitle{Fairness Through Unawareness}
    The most straightforward solution to fairness seems to be that
    just simply dropping all the protected columns.
    \begin{itemize}
        \item This is called fairness through unawareness.
        \item Formally it's
        \[
            X_i = X_j \implies \hat{Y}_i = \hat{Y}_j
        \]
        where $i,j$ are individuals; $X$ is the set of attributes
        except protected attributes; and $\hat{Y}$ is the prediction.
        \item Also known as fairness through blindness and
        anti-classification.
    \end{itemize}
\end{frame}

\begin{frame}
    \frametitle{Fairness Through Unawareness}
    The downside of this is there could still be ``proxy'' attributes
    that correlate with protected attributes: like Occupation still
    correlates with Income.
    \begin{table}
        \begin{tabular}{|c|c|c|c|c|}
            \hline
            \sout{Age} & \sout{Gender}   & \red{Occupation}  & \red{\sout{Income}}   & Education \\
            \hline
            \sout{28} & \sout{M} & \red{Engineering} & \red{\sout{\$80,000}} & Master \\
            \sout{28} & \sout{F} & \red{Engineering} & \red{\sout{\$65,000}} & Master \\
            \sout{45} & \sout{M} & \red{Medicine}    & \red{\sout{\$100,000}} & Doctorate \\
            \sout{40} & \sout{F} & \red{Legal}       & \red{\sout{\$150,000}} & Law Degree \\
            \sout{32} & \sout{M} & \red{Education}   & \red{\sout{\$55,000}} & Bachelor \\
            \hline
        \end{tabular}
        \caption{Example Dataset}
    \end{table}
\end{frame}

\begin{frame}
    \frametitle{Demographic Parity}
    Formally, it requires that
    \[
        \abs{P[\hat{Y} = 1 | S = 1] - P[\hat{Y} = 1 | S \neq 1]} \leq \epsilon
    \]
    where $\hat{Y} = 1$ represents acceptance(positive);
    $S = 1$ represents privileged group;
    $S \neq 1$ represents unprivileged group where $S$ is some protected
    attributes.

    Let's set $\epsilon = 0.2$.
    If for some job opening there are $10$ female applicants and
    $100$ male applicants, and there are $8$ accepted females
    and $90$ accepted males:
    \[
        \abs{8/10 - 90/100} = 0.1 < \epsilon \quad\text{so this is fair}
    \]
    while if there were $1$ accepted females and $50$ accepted males:
    \[
        \abs{1/10 - 50/100} = 0.4 > \epsilon \quad\text{so this is unfair}
    \]
\end{frame}

\begin{frame}
    \frametitle{Disadvantages of Demographic Parity}
    \begin{itemize}
        \item A fully accurate classifier
        may be considered unfair.
        \item The notion permits that we accept the qualified applicants in
        one demographic, but random individuals in another, so long as the
        percentages of acceptance match.
        \item For example, the case with 90 qualified males and only 1
        qualified females.
    \end{itemize}
\end{frame}

\begin{frame}
    \frametitle{Equalized Odds}
    \begin{itemize}
        \item Equalized odds is designed to address the downsides of the previous two
        by taking into accounts the ``ground truths'' and consider
        the difference between false-positive rates and true-positive rates of
        the groups.
        \item Formally, it requires that
        \begin{align*}
            \abs{P[\hat{Y} = 1 | S = 1, Y = 0] - P[\hat{Y} = 1 | S \neq 1, Y = 0]} & \leq \epsilon \\
            \abs{P[\hat{Y} = 1 | S = 1, Y = 1] - P[\hat{Y} = 1 | S \neq 1, Y = 1]} & \leq \epsilon
        \end{align*}
        where $Y$ represents ground truths.
        \item A fully accurate classifier
        will necessarily satisfy the two equalized odds constraints.
    \end{itemize}
\end{frame}

\begin{frame}
    \frametitle{Equal Opportunity}
    \begin{itemize}
        \item It is a relaxation of equalized odds.
        \item Formally, it requires that
        \begin{align*}
            \abs{P[\hat{Y} = 1 | S = 1, Y = 1] - P[\hat{Y} = 1 | S \neq 1, Y = 1]} & \leq \epsilon
        \end{align*}
        where $Y$ represents ground truths.
    \end{itemize}
\end{frame}

\begin{frame}
    \frametitle{COMPAS}
    \textbf{recidivism} \textit{noun} \\
    \quad the tendency of a convicted criminal to reoffend.
    \begin{columns}[T]
        \begin{column}{.5\textwidth}
            \centering
            \includegraphics[width=.5\textwidth]{PRATER.jpg}
            \footnotesize
            \begin{itemize}
                \item Prior Offenses: 2 armed robberies, 1 attempted armed robbery
                \item Subsequent Offenses: 1 grand theft
                \item Risk Score: 3
            \end{itemize}
        \end{column}
        \begin{column}{.5\textwidth}
            \centering
            \includegraphics[width=.5\textwidth]{BORDEN.jpg}
            \footnotesize
            \begin{itemize}
                \item Prior Offenses: 4 juvenile misdemeanors
                \item Subsequent Offenses: None
                \item Risk Score: 8
            \end{itemize}
        \end{column}
    \end{columns}
\end{frame}

\begin{frame}
    \frametitle{COMPAS}
    \begin{itemize}
        \item COMPAS is an algorithm
        used by U.S. courts for predicting recidivism based on a
        questionaire and background information.
        \item In 2016, ProPublica found that the algorithm is biased.
        \begin{quote}
            Black defendants were often predicted to be at a higher risk of recidivism than they actually were.
            White defendants were often predicted to be less risky than they were.
        \end{quote}
        \item The false-positive rates vary significantly across
        black people and white people, violating equalized odds.
        \item Supreme Court ruled that it can be considered by judges during
        sentencing, but there must be warnings about the tool's "limitations and cautions."
    \end{itemize}
    \footnotetext[1]{\href{https://www.propublica.org/article/how-we-analyzed-the-compas-recidivism-algorithm}{(Link) ProPublica - How We Analyzed the COMPAS Recidivism Algorithm}}
    \footnotetext[2]{\href{https://www.youtube.com/watch?v=TqnYf2h6A-k}{(Link) Vsauce2 - The Dangerous Math Used To Predict Criminals}}
\end{frame}

\begin{frame}
    \frametitle{Other Measures}
    \begin{itemize}
        \item Overall accuracy equality:
        \[
            \abs{P[Y = \hat{Y} | S = 1] - P[Y = \hat{Y} | S \neq 1]} \leq \epsilon
        \]
        where $Y = \hat{Y}$ means that the prediction was correct.
        \item Predictive parity:
        \[
            \abs{P[Y = 1 | S = 1, \hat{Y} = 1] - P[Y = 1 | S \neq 1, \hat{Y} = 1]} \leq \epsilon
        \]
        This requires that the ``positive predictive values'' are similar across
        groups, meaning the probability of an individual with a positive prediction
        actually experiencing a positive outcome.
    \end{itemize}
\end{frame}

\begin{frame}
    \frametitle{Framework}
    We made a tool to calculate these measures.

    Our framework consists of 3 parties: data provider, model maker, and auditor.
    \begin{itemize}
        \item Data provider has access to the real world data.
        For example, a census bureau.
        \item Model maker designs AI models that is to be used on data.
        They can use our tool on their in-house data to test their models.
        \item Auditor is some 3rd-party that takes real data and a model or
        model result.
        They can use our tool to determine the fairness of the model.
    \end{itemize}
\end{frame}

\newcommand{\db}{\mathcal{D}}
\newcommand{\model}{\mathcal{M}}
\newcommand{\priv}{R}
\newcommand{\pos}{\hat{P}}
\newcommand{\posm}{\hat{P}_\mathcal{M}}
\newcommand{\tru}{T}
\newcommand{\leg}{L}
\newcommand{\sco}{\hat{S}}
\newcommand{\scom}{\hat{S}_\mathcal{M}}
\newcommand{\calib}{C}

\begin{frame}
    \frametitle{Abstraction}
    Our framework tool abstracts the fairness measures:
    \begin{itemize}
        \item A \emph{row} is a lookup table or dictionary.
        $r_n(\text{``sex''}) = \text{``Female''}$ means $r_n$'s sex is female.
        \item A \emph{privileged predicate} $\priv$ takes a row and determines if
        it belongs to the privileged group.
        For example, $\priv(r_i) := r_i(\text{``race''}) == \text{``Caucasian''}$
        means the privileged group is those with race being Caucasian.
        \item A \emph{positive predicate} $\pos$ takes a row and determines if its
        prediction is positive. For example,
        $\pos(r_i) := \mathtt{int}(r_i(\text{``score''})) > 7$ means a
        row's prediction is positive if its score is greater than 7.
        \item A \emph{ground truth predicate} $\tru$ takes a row and gives the ground
        truth of the result. For example,
        $\tru(r_i) := r_i(\text{``recid''})$ indicates a
        row's actual recidivism.
    \end{itemize}
\end{frame}

\begin{frame}
    \frametitle{Definition}
    We can use these abstractions to define fairness measures.
    For equal opportunity, recall its formal definition:
    \begin{align*}
        \abs{P[\hat{Y} = 1 | S = 1, Y = 1] - P[\hat{Y} = 1 | S \neq 1, Y = 1]}
        \leq \epsilon
    \end{align*}
    To model $Y$, $\hat{Y}$, and $S$, we define the corresponding
    $\tru$, $\pos$, and $\priv$.
    \begin{itemize}
        \item $Y = 1$ if and only if $\tru$ is true
        \item $\hat{Y} = 1$ if and only if $\pos$ is true
        \item $S = 1$ if and only if $\priv$ is true
    \end{itemize}
    This way, we can calculate equal opportunity as a function:
    \begin{align*}
        \text{equal\_opportunity} (\epsilon, \model, \priv, \pos, \tru)
    \end{align*}
\end{frame}

\begin{frame}
    \frametitle{Application}
    To demonstrate an application, we analyzed the COMPAS dataset.
    We set the predicates:
    \begin{align*}
        \priv(r_i) & := r_i(\text{``race''}) \neq \text{``African-American''}\\
        \pos(r_i)  & := r_i(\text{``score\_text''}) \in \set{\text{``Medium''},\text{``High''}}\\
        \tru(r_i)  & := r_i(\text{``two\_year\_recid''}) == \text{True}
    \end{align*}
    These predicates can be published for transparency.
\end{frame}

\begin{frame}
    \frametitle{Application}
    \centering
    \includegraphics[width=0.73\textwidth]{African-American}
    \captionof{figure}{Unprivileged Group: African-American}
\end{frame}

\begin{frame}
    \frametitle{Application}
    \centering
    \includegraphics[width=0.73\textwidth]{Caucasian}
    \captionof{figure}{Unprivileged Group: Caucasian}
\end{frame}

\begin{frame}
    \frametitle{Conclusion}
    \begin{itemize}
        \item Decision making by AI may be biased.
        \item With our framework tool, auditors can comprehensively
        review the fairness of an AI system.
    \end{itemize}
\end{frame}

\end{document}

### 2024-08-17 slides/2024_8_30.tex
Updated slides

--- 2024-08-16 slides/2024_8_20.tex
@@
 
-\newdate{date}{20}{08}{2024}
+\newdate{date}{30}{08}{2024}
 \date{\displaydate{date}}
@@
 \begin{frame}
-    \frametitle{Bias}
+    \frametitle{AI Bias}
     \begin{itemize}
@@
     Let's set $\epsilon = 0.2$.
-    If for some job opening there are $10$ female applicants and
+    If for some job opening, male is the privileged group and
+    there are $10$ female applicants and
     $100$ male applicants, and there are $8$ accepted females
@@
     \frametitle{Other Measures}
-    \begin{itemize}
-        \item Overall accuracy equality:
-        \[
-            \abs{P[Y = \hat{Y} | S = 1] - P[Y = \hat{Y} | S \neq 1]} \leq \epsilon
-        \]
-        where $Y = \hat{Y}$ means that the prediction was correct.
-        \item Predictive parity:
-        \[
-            \abs{P[Y = 1 | S = 1, \hat{Y} = 1] - P[Y = 1 | S \neq 1, \hat{Y} = 1]} \leq \epsilon
-        \]
-        This requires that the ``positive predictive values'' are similar across
-        groups, meaning the probability of an individual with a positive prediction
-        actually experiencing a positive outcome.
-    \end{itemize}
+        \begin{table}[h]
+            \centering
+            \resizebox{0.93\textwidth}{!}{
+            % \resizebox{\columnwidth}{!}{
+            \begin{tabular}{|l|c|} % 'l' specifies left alignment for the first column
+                \hline
+                \textbf{Fairness Measure} & \textbf{Definition} \\
+                \hline
+        Disparate Impact & $\frac{P[\hat{Y} = 1 | S \neq 1]}{P[\hat{Y} = 1 | S = 1]} \geq 1 - \epsilon$ \\
+        \hline
+        Demographic Parity & $\abs{P[\hat{Y} = 1 | S = 1] - P[\hat{Y} = 1 | S \neq 1]} \leq \epsilon$ \\
+        \hline
+        Conditional Statistical Parity & $\abs{P[\hat{Y} = 1 | S = 1, L = l] - P[\hat{Y} = 1 | S \neq 1, L = l]} \leq \epsilon$ \\
+        \hline
+        Overall Accuracy Equality & $\abs{P[Y = \hat{Y} | S = 1] - P[Y = \hat{Y} | S \neq 1]} \leq \epsilon$ \\
+        \hline
+        Mean Difference & $\abs{E[\hat{Y}|S = 1] - E[\hat{Y}|S \neq 1]} \leq \epsilon$ \\
+        \hline
+        \multirow{2}{*}{Equalized Odds} & $\abs{P[\hat{Y} = 1 | S = 1, Y = 0] - P[\hat{Y} = 1 | S \neq 1, Y = 0]} \leq \epsilon$ \\
+        \cline{2-2}
+                                        & $\abs{P[\hat{Y} = 1 | S = 1, Y = 1] - P[\hat{Y} = 1 | S \neq 1, Y = 1]} \leq \epsilon$ \\
+        \hline
+        Equal Opportunity & $\abs{P[\hat{Y} = 1 | S = 1, Y = 1] - P[\hat{Y} = 1 | S \neq 1, Y = 1]} \leq \epsilon$ \\
+        \hline
+        Predictive Equality & $\abs{P[\hat{Y} = 1 | S = 1, Y = 0] - P[\hat{Y} = 1 | S \neq 1, Y = 0]} \leq \epsilon$ \\
+        \hline
+        \multirow{2}{*}{Conditional Use Accuracy Equality} & $\abs{P[Y = 1 | S = 1, \hat{Y} = 1] - P[Y = 1 | S \neq 1, \hat{Y} = 1]} \leq \epsilon$ \\
+        \cline{2-2}
+                                         & $\abs{P[Y = 0 | S = 1, \hat{Y} = 0] - P[Y = 0 | S \neq 1, \hat{Y} = 0]} \leq \epsilon$ \\
+        \hline
+        Predictive Parity & $\abs{P[Y = 1 | S = 1, \hat{Y} = 1] - P[Y = 1 | S \neq 1, \hat{Y} = 1]} \leq \epsilon$ \\
+        \hline
+        Equal Calibration & $\abs{P[Y = 1 | S = 1, \hat{V} = v] - P[Y = 1 | S \neq 1, \hat{V} = v]} \leq \epsilon$ \\
+        \hline
+        Positive Balance & $\abs{E[\hat{V} | Y = 1, S = 1] - E[\hat{V} | Y = 1, S \neq 1]} \leq \epsilon$ \\
+        \hline
+        Negative Balance & $\abs{E[\hat{V} | Y = 0, S = 1] - E[\hat{V} | Y = 0, S \neq 1]} \leq \epsilon$ \\
+        \hline
+            \end{tabular}
+            }
+            \caption{Fairness measures.}
+            \label{tab:measures}
+        \end{table}
 \end{frame}
@@
 \begin{frame}
+    \frametitle{Predicate}
+    We first introduce the idea of a predicate.
+    \begin{itemize}
+        \item A \emph{predicate} is a decider that determines
+        if some condition holds for some given input.
+        \item Formally, it is a function that takes
+        some type $\alpha$ and returns True or False.
+        \item For example, an ``$\text{adult}(x)$'' predicate might
+        take a person $x$'s data and returns if they are over 18
+        year old.
+    \end{itemize}
+\end{frame}
+
+\begin{frame}
     \frametitle{Abstraction}
@@
         truth of the result. For example,
-        $\tru(r_i) := r_i(\text{``recid''})$ indicates a
+        $\tru(r_i) := r_i(\text{``recid''}) == \text{True}$ indicates a
         row's actual recidivism.
@@
         \item $S = 1$ if and only if $\priv$ is true
+        \item and vice versa
     \end{itemize}
@@
     \begin{align*}
-        \text{equal\_opportunity} (\epsilon, \model, \priv, \pos, \tru)
+        \text{equal\_opportunity} (\epsilon, \priv, \pos, \tru)
     \end{align*}
@@
     \frametitle{Application}
-    To demonstrate an application, we analyzed the COMPAS dataset.
-    We set the predicates:
+    For demonstration, we analyzed the COMPAS dataset.
+    Let's set the predicates:
     \begin{align*}
@@
 \begin{frame}
+    \frametitle{Application}
+    \centering
+    \includegraphics[width=0.73\textwidth]{Asian}
+    \captionof{figure}{Unprivileged Group: Asian (32 rows)}
+\end{frame}
+
+\begin{frame}
     \frametitle{Conclusion}

### 2024-08-17 slides/2024_8_30.tex
Added code snippets

@@
 \usepackage{subcaption}
+\usepackage{minted}
 \usepackage{xcolor}
@@
 
+\begin{frame}[fragile]
+    \frametitle{Definition}
+    \begin{align*}
+        \text{equal\_opportunity} (\epsilon, \priv, \pos, \tru)
+    \end{align*}
+
+    This is implemented in our tool as a Python function:
+
+    \begin{center}
+    \begin{minipage}{0.8\textwidth}
+    \begin{minted}[fontsize=\small, frame=single, linenos]{python}
+def equal_opportunity(
+    ratio,
+    privileged_predicate,
+    positive_predicate,
+    truth_predicate):
+    ...
+    \end{minted}
+    \end{minipage}
+    \end{center}
+\end{frame}
+
 \begin{frame}
@@
     These predicates can be published for transparency.
+\end{frame}
+
+\begin{frame}[fragile]
+    \frametitle{Application}
+    \begin{align*}
+        \priv(r_i) & := r_i(\text{``race''}) \neq \text{``African-American''}\\
+        \pos(r_i)  & := r_i(\text{``score\_text''}) \in \set{\text{``Medium''},\text{``High''}}\\
+        \tru(r_i)  & := r_i(\text{``two\_year\_recid''}) == \text{True}
+    \end{align*}
+    These predicates can then be encoded with Python functions:
+    \begin{center}
+    \begin{minipage}{0.8\textwidth}
+    \begin{minted}[fontsize=\small, frame=single, linenos]{python}
+equal_opportunity(
+    0.2,
+    lambda row: row['race'] == 'African-American',
+    lambda row: row['score_text']
+      in {'Medium', 'High'},
+    lambda row: row['is_recid'] == '1')
+    \end{minted}
+    \end{minipage}
+    \end{center}
 \end{frame}
@@
     \end{itemize}
+
+    This slides is available at \url{https://github.com/RexYuan/Shu}.
 \end{frame}

### 2024-08-18 slides/2024_8_30.tex
Fixed typo

@@
     0.2,
-    lambda row: row['race'] == 'African-American',
+    lambda row: row['race'] != 'African-American',
     lambda row: row['score_text']

### 2024-08-21 setup.py
Fixed error

@@ -5,7 +5,7 @@ with open("src/README.md", "r") as f:
 
 setup(
     name="fairness_checker",
-    version="0.1.2",
+    version="0.1.3",
     package_dir={'': 'src'},
     packages=find_packages(where="src"),
     description="Fairnes checker",

### 2024-08-21 src/fairness_checker/main.py

@@ -33,10 +33,10 @@ class fairness_model_checker:
 
     def write_to_csv(self, filename: str, rows: Iterable[csv_row]):
         with open(filename, 'w') as csvfile:
-            writer = csv.DictWriter(csvfile, fieldnames=self.fieldnames[:-1])
+            writer = csv.DictWriter(csvfile, fieldnames=self.fieldnames)
             writer.writeheader()
             for row in rows:
-                writer.writerow({k: v for k, v in row.items() if k in self.fieldnames[:-1]})
+                writer.writerow({k: v for k, v in row.items() if k in self.fieldnames})
 
     def remove_csv(self, filename: str):
         os.remove(filename)

### 2024-08-23 setup.py
Fixed typing

@@ -5,7 +5,7 @@ with open("src/README.md", "r") as f:
 
 setup(
     name="fairness_checker",
-    version="0.1.3",
+    version="0.1.11",
     package_dir={'': 'src'},
     packages=find_packages(where="src"),
     description="Fairnes checker",
@@ -15,10 +15,13 @@ setup(
     author="RexYuan",
     author_email="hello@rexyuan.com",
     license="Unlicense",
-    python_requires=">=3.6",
+    python_requires=">=3.8",
     entry_points={
         'console_scripts': [
             'fairness_checker=fairness_checker.__main__:main'
         ]
+    },
+    package_data={
+        'fairness_checker': ['py.typed'],
     },
 )

### 2024-08-23 src/fairness_checker/main.py

@@ -2,11 +2,15 @@ import os
 import csv
 import math
 
-from typing import Callable, Iterable, Dict, TypeVar, Any, Tuple
+from typing import Callable, Iterable, Dict, TypeVar, Any, Tuple, Protocol, Union, Sequence
 from contextlib import contextmanager
 
 csv_row = Dict[str, str]
 T = TypeVar('T')
+K = TypeVar('K')
+class Predictable(Protocol):
+    def predict(self, filename : str) -> Sequence[Any]:
+        ...
 
 def print_stats(title, measure, ratio):
     print(title)
@@ -25,10 +29,10 @@ class fairness_model_checker:
         Parameters:
         raw_file (str): Path to the raw data file in CSV format.
         """
-        self.verbose = verbose
+        self.verbose: bool = verbose
         with open(raw_file, 'r') as file:
             reader = csv.DictReader(file)
-            self.fieldnames = reader.fieldnames
+            self.fieldnames = list(reader.fieldnames or [])
             self.reader = list(reader)
 
     def write_to_csv(self, filename: str, rows: Iterable[csv_row]):
@@ -49,10 +53,10 @@ class fairness_model_checker:
 
     def disparate_impact(self,
                          ratio: float,
-                         model: object,
+                         model: Predictable,
                          privileged_predicate: Callable[[csv_row], bool],
                          positive_predicate: Callable[[T], bool],
-                         value: bool = False) -> bool:
+                         value: bool = False) -> Union[bool, float]:
         """
         Evaluates the disparate impact of the model's predictions between privileged and unprivileged groups.
 
@@ -88,10 +92,10 @@ class fairness_model_checker:
 
     def demographic_parity(self,
                            ratio: float,
-                           model: object,
+                           model: Predictable,
                            privileged_predicate: Callable[[csv_row], bool],
                            positive_predicate: Callable[[T], bool],
-                           value: bool = False) -> bool:
+                           value: bool = False) -> Union[bool, float]:
         """
         Evaluates the demographic parity of the model's predictions between privileged and unprivileged groups.
 
@@ -127,11 +131,11 @@ class fairness_model_checker:
 
     def equalized_odds(self,
                        ratio: float,
-                       model: object,
+                       model: Predictable,
                        privileged_predicate: Callable[[csv_row], bool],
                        positive_predicate: Callable[[T], bool],
                        truth_predicate: Callable[[csv_row], bool],
-                       value: bool = False) -> bool:
[…]

### 2024-08-23 src/test/test_model.py

@@ -113,16 +113,24 @@ trained_tree = compas_model_wrapper(tree_model)
 
 c = fairness_model_checker('compas_score.csv')
 race = 'African-American'
-c.disparate_impact(0.8, trained, lambda row: row['race'] != race, lambda Y: Y == 1)
-c.demographic_parity(0.2, trained, lambda row: row['race'] != race, lambda Y: Y == 1)
-c.equalized_odds(0.2, trained, lambda row: row['race'] != race, lambda Y: Y == 1, lambda row: row['two_year_recid'] == '1')
-c.equal_opportunity(0.2, trained, lambda row: row['race'] != race, lambda Y: Y == 1, lambda row: row['two_year_recid'] == '1')
-c.accuracy_eqaulity(0.2, trained, lambda row: row['race'] != race, lambda Y: Y == 1, lambda row: row['is_recid'] == '1')
-c.predictive_parity(0.2, trained, lambda row: row['race'] != race, lambda Y: Y == 1, lambda row: row['is_recid'] == '1')
-c.equal_calibration(0.2, trained_tree, lambda row: row['race'] != race, lambda row: row['is_recid'] == '1', lambda u,l: (lambda Y: l <= int(Y) and int(Y) <= u), (7, 5))
-c.conditional_statistical_parity(0.2, trained, lambda row: row['race'] != race, lambda Y: Y == 1, lambda x: (lambda row: int(row['priors_count']) > x), (0,))
-c.predictive_equality(0.2, trained, lambda row: row['race'] != race, lambda Y: Y == 1, lambda row: row['is_recid'] == '1')
-c.conditional_use_accuracy_equality(0.2, trained, lambda row: row['race'] != race, lambda Y: Y == 1, lambda row: row['is_recid'] == '1')
-c.positive_balance(5, trained_tree, lambda row: row['race'] != race, lambda Y: int(Y), lambda row: row['is_recid'] == '1')
-c.negative_balance(5, trained_tree, lambda row: row['race'] != race, lambda Y: int(Y), lambda row: row['is_recid'] == '1')
-c.mean_difference(0.2, trained, lambda row: row['race'] != race, lambda Y: Y == 1)
+def pos_pred(Y: int) -> bool:
+    return Y == 1
+def calib_pred(u: int, l: int):
+    def pred(Y: int) -> bool:
+        return l <= int(Y) and int(Y) <= u
+    return pred
+def sco_pred(Y: int) -> int:
+    return int(Y)
[…]

### 2024-08-24 setup.py
Added more description

@@ -5,7 +5,7 @@ with open("src/README.md", "r") as f:
 
 setup(
     name="fairness_checker",
-    version="0.1.11",
+    version="0.1.12",
     package_dir={'': 'src'},
     packages=find_packages(where="src"),
     description="Fairnes checker",

### 2024-08-29 slides/2024_8_30.tex
Added more description
Added slides

@@
         if some condition holds for some given input.
-        \item Formally, it is a function that takes
-        some type $\alpha$ and returns True or False.
         \item For example, an ``$\text{adult}(x)$'' predicate might
@@
     \end{itemize}
+\end{frame}
+
+\begin{frame}
+    \frametitle{Predicate}
+    \[
+    \begin{array}{rcl}
+    \text{adult}(r_i) & := & r_i(\text{``Age''}) > 18 \\
+    \text{rich}(r_i) & := & r_i(\text{``Income''}) > 3000
+    \end{array}
+    \]
+    \begin{table}
+        \begin{tabular}{|c|c|c|c|c|}
+            \hline
+                  & Age & Income & $\text{adult}(r_i)$ & $\text{rich}(x)$ \\
+            \hline
+            $r_1$ & 14 & \$10000 & False & True  \\
+            $r_2$ & 17 & \$1000 & False & False  \\
+            $r_3$ & 45 & \$1500 & True & False  \\
+            $r_4$ & 40 & \$2000 & True & False  \\
+            $r_5$ & 32 & \$5000 & True & True  \\
+            \hline
+        \end{tabular}
+        \caption{Example Dataset with Predicates}
+    \end{table}
 \end{frame}
@@
         prediction is positive. For example,
-        $\pos(r_i) := \mathtt{int}(r_i(\text{``score''})) > 7$ means a
+        $\pos(r_i) := r_i(\text{``score''}) > 7$ means a
         row's prediction is positive if its score is greater than 7.
@@
 
-    This slides is available at \url{https://github.com/RexYuan/Shu}.
+    Slides: \url{https://github.com/RexYuan/Shu}.
+
+    Package: \url{https://pypi.org/project/fairness-checker}.
 \end{frame}

### 2024-12-12 slides/2024_12_13.tex
Added Dec 13rd slides

--- 2024-08-29 slides/2024_8_30.tex
@@
 
-\newdate{date}{30}{08}{2024}
+\newdate{date}{13}{12}{2024}
 \date{\displaydate{date}}
-\title[Fairness]{Ensuring Fairness with Transparent Auditing of
+\title[AI Fairness]{Ensuring Fairness with Transparent Auditing of
 Quantitative Bias in AI Systems}
@@
         it belongs to the privileged group.
-        For example, $\priv(r_i) := r_i(\text{``race''}) == \text{``Caucasian''}$
+        For example, $\priv(r_i) := r_i(\text{``race''}) = \text{``Caucasian''}$
         means the privileged group is those with race being Caucasian.
@@
         truth of the result. For example,
-        $\tru(r_i) := r_i(\text{``recid''}) == \text{True}$ indicates a
+        $\tru(r_i) := r_i(\text{``recid''}) = \text{True}$ indicates a
         row's actual recidivism.
@@
 
-\begin{frame}[fragile]
-    \frametitle{Definition}
-    \begin{align*}
-        \text{equal\_opportunity} (\epsilon, \priv, \pos, \tru)
-    \end{align*}
-
-    This is implemented in our tool as a Python function:
-
-    \begin{center}
-    \begin{minipage}{0.8\textwidth}
-    \begin{minted}[fontsize=\small, frame=single, linenos]{python}
-def equal_opportunity(
-    ratio,
-    privileged_predicate,
-    positive_predicate,
-    truth_predicate):
-    ...
-    \end{minted}
-    \end{minipage}
-    \end{center}
-\end{frame}
-
 \begin{frame}
@@
         \pos(r_i)  & := r_i(\text{``score\_text''}) \in \set{\text{``Medium''},\text{``High''}}\\
-        \tru(r_i)  & := r_i(\text{``two\_year\_recid''}) == \text{True}
+        \tru(r_i)  & := r_i(\text{``two\_year\_recid''}) = \text{True}
     \end{align*}
     These predicates can be published for transparency.
-\end{frame}
-
-\begin{frame}[fragile]
-    \frametitle{Application}
-    \begin{align*}
-        \priv(r_i) & := r_i(\text{``race''}) \neq \text{``African-American''}\\
-        \pos(r_i)  & := r_i(\text{``score\_text''}) \in \set{\text{``Medium''},\text{``High''}}\\
-        \tru(r_i)  & := r_i(\text{``two\_year\_recid''}) == \text{True}
-    \end{align*}
-    These predicates can then be encoded with Python functions:
-    \begin{center}
-    \begin{minipage}{0.8\textwidth}
-    \begin{minted}[fontsize=\small, frame=single, linenos]{python}
-equal_opportunity(
-    0.2,
-    lambda row: row['race'] != 'African-American',
-    lambda row: row['score_text']
-      in {'Medium', 'High'},
-    lambda row: row['is_recid'] == '1')
-    \end{minted}
-    \end{minipage}
-    \end{center}
 \end{frame}
@@
 \begin{frame}
-    \frametitle{Application}
-    \centering
-    \includegraphics[width=0.73\textwidth]{Asian}
-    \captionof{figure}{Unprivileged Group: Asian (32 rows)}
-\end{frame}
-
-\begin{frame}
     \frametitle{Conclusion}
@@
         \item Decision making by AI may be biased.
-        \item With our framework tool, auditors can comprehensively
+        \item With our framework and tool, auditors can comprehensively
         review the fairness of an AI system.
